---
title: "Peve enrichment of miRNA targets"
author: "Jill Ashey"
date: "2024-10-30"
output: html_document
---

## GO enrichment of Peve miRNA targets

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(topGO)
```

See [bash code](https://github.com/JillAshey/JillAshey_Putnam_Lab_Notebook/blob/master/_posts/2024-06-15-e5-deepdive-miRNA-TargetPrediction.md) for `miranda` analysis, which was done on Andromeda. 

Read in miranda output file 
```{r}
miranda_1kb <- read.delim("../output/21-Peve-miRNA-target-prediction/miranda_strict_all_1kb_parsed_peve.txt", header = F) %>% 
  rename(mirna = V1,
        Target = V2, 
        Score = V3, 
        Energy_kcal_mol = V4, 
        Query_start_end = V5, 
        Subject_start_end = V6,
        Length = V7, 
        Subject_Identity = V8, 
        Query_Identity = V9)

# Remove > in front of miRNA name 
miranda_1kb$mirna <- gsub(">", "", miranda_1kb$mirna)
length(unique(miranda_1kb$mirna))
length(unique(miranda_1kb$Target))
```

Lost one miRNA in miranda cutoffs. 

Make gene id column
```{r}
miranda_1kb$gene_id <- sub("^ID=([^;]+);.*", "\\1", miranda_1kb$Target)
```

Create new df with only gene and miRNA info; summarize which miRNAs target which genes 
```{r}
# Create the new data frame with gene_id and concatenated mirnas
gene_mirna_1kb_df <- miranda_1kb %>%
  group_by(gene_id) %>%
  summarize(mirnas = paste(mirna, collapse = ", "))
```

Summarize number of genes targeted by each miRNA
```{r}
mirna_gene_summary <- miranda_1kb %>%
  group_by(mirna) %>%
  summarize(num_genes = n_distinct(gene_id))
```

Read in gene to go information (generated [here](https://github.com/urol-e5/deep-dive/blob/main/E-Peve/code/19-Peve-bedtools-closest.Rmd))
```{r}
PEVE_gene2go<-read.delim("../../data/Peve_gene2go.tab", sep="\t")
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
peve_clust_genes <- as.character(unique(miranda_1kb$gene_id))

# All genes 
peve_all_genes <- as.character(PEVE_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
Peve_GeneList <- factor(as.integer(peve_all_genes %in% peve_clust_genes))
names(Peve_GeneList) <- peve_all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes closest to the piRNA clusters. 

Read in gene-to-go-mappings
```{r}
PEVE_gene2go_topgo<-readMappings("../../data/Peve_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_PEVE_BP <-new("topGOdata", ontology="BP", gene2GO=PEVE_gene2go_topgo, allGenes=Peve_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_PEVE_BP_FE <- runTest(GO_PEVE_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_PEVE_BP_En <- GenTable(GO_PEVE_BP, Fisher = GO_PEVE_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_PEVE_BP_En$Fisher<-as.numeric(GO_PEVE_BP_En$Fisher)
GO_PEVE_BP_En_sig<-GO_PEVE_BP_En[GO_PEVE_BP_En$Fisher<0.05,]
```

Merge `GO_PEVE_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
PEVE_gene2go <- PEVE_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
PEVE_gene2go$GeneOntologyIDs <- trimws(PEVE_gene2go$GeneOntologyIDs)
GO_PEVE_BP_En_sig$GO.ID <- trimws(GO_PEVE_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_PEVE_BP_En_sig_gene <- PEVE_gene2go %>%
  left_join(GO_PEVE_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_PEVE_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_PEVE_CC <-new("topGOdata", ontology="CC", gene2GO=PEVE_gene2go_topgo, allGenes=Peve_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_PEVE_CC_FE <- runTest(GO_PEVE_CC, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_PEVE_CC_En <- GenTable(GO_PEVE_CC, Fisher = GO_PEVE_CC_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_PEVE_CC_En$Fisher<-as.numeric(GO_PEVE_CC_En$Fisher)
GO_PEVE_CC_En_sig<-GO_PEVE_CC_En[GO_PEVE_CC_En$Fisher<0.05,]
```

Merge `GO_PEVE_CC_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
PEVE_gene2go <- PEVE_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
PEVE_gene2go$GeneOntologyIDs <- trimws(PEVE_gene2go$GeneOntologyIDs)
GO_PEVE_CC_En_sig$GO.ID <- trimws(GO_PEVE_CC_En_sig$GO.ID)

# Join the datasets based on GO term
GO_PEVE_CC_En_sig_gene <- PEVE_gene2go %>%
  left_join(GO_PEVE_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_PEVE_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions  

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_PEVE_MF <-new("topGOdata", ontology="MF", gene2GO=PEVE_gene2go_topgo, allGenes=Peve_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_PEVE_MF_FE <- runTest(GO_PEVE_MF, algorithm="weight01", statistic="fisher")
```


Generate results table 
```{r}
GO_PEVE_MF_En <- GenTable(GO_PEVE_MF, Fisher = GO_PEVE_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_PEVE_MF_En$Fisher<-as.numeric(GO_PEVE_MF_En$Fisher)
GO_PEVE_MF_En_sig<-GO_PEVE_MF_En[GO_PEVE_MF_En$Fisher<0.05,]
```

Merge `GO_PEVE_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
PEVE_gene2go <- PEVE_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
PEVE_gene2go$GeneOntologyIDs <- trimws(PEVE_gene2go$GeneOntologyIDs)
GO_PEVE_MF_En_sig$GO.ID <- trimws(GO_PEVE_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_PEVE_MF_En_sig_gene <- PEVE_gene2go %>%
  left_join(GO_PEVE_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_PEVE_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_PEVE_En_sig_gene <- rbind(GO_PEVE_BP_En_sig_gene, GO_PEVE_CC_En_sig_gene, GO_PEVE_MF_En_sig_gene)
```

Merge `miranda_1kb` with `GO_PEVE_En_sig_gene` and calculate proportion of significant genes
```{r}
merged_targets <- GO_PEVE_En_sig_gene %>%
  left_join(miranda_1kb, by = c("gene_name" = "gene_id")) %>%
  na.omit() %>%
  mutate(prop.sig.genes = Significant/Annotated)

# Write to csv 
write.csv(merged_targets, "../output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_miRNA_targets.csv")
```

Plot 
```{r}
PEVE_GO_plot<-ggplot(merged_targets, aes(x = Term, y = Fisher, size = prop.sig.genes, fill = Fisher)) +
  #expand_limits(y = 1.5) +
  #ylim(1, 7.25) +
  # Add horizontal lines with a single aesthetic value
  #geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  #geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  #labs(caption = 'Cut-off lines at p=0.01 and p=0.001') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); PEVE_GO_plot
```

