---
title: "10-shortRNA-ShortStack-comparison"
author: "Kathleen Durkin"
date: "2024-05-28"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(DT)
library(Biostrings)
library(tm)
library(ggvenn)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center", # Align plots to the center
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

I want to find miRNAs that are conserved among either a subset of or all three species of interest (*A.pulchra*, *P.evermanni*, and *P.meandrina*) using Blastn. 

# Prep data

## Isolate mature miRNA sequences

Our ShortStack output contains sequences for the mature, star, and precursor sequences for each identified miRNA. We just want to look at the mature miRNA sequences right now, so let's isolate those.

```{r, engine='bash'}
cd ../data/10-shortRNA-ShortStack-comparison

# Copy all sequences whose headers contain "mature"
awk '/^>/ {p = /mature/} p' ../../../D-Apul/output/13.2.1-Apul-sRNAseq-ShortStack-31bp-fastp-merged-cnidarian_miRBase/ShortStack_out/mir.fasta > Apul_ShortStack_mature.fasta

awk '/^>/ {p = /mature/} p' ../../../E-Peve/output/08.2-Peve-sRNAseq-ShortStack-31bp-fastp-merged/ShortStack_out/mir.fasta > Peve_ShortStack_mature.fasta

awk '/^>/ {p = /mature/} p' ../../../F-Pmea/output/13.2.1-Pmea-sRNAseq-ShortStack-31bp-fastp-merged-cnidarian_miRBase/ShortStack_out/mir.fasta > Pmea_ShortStack_mature.fasta

grep "^>" Apul_ShortStack_mature.fasta | wc -l 
grep "^>" Peve_ShortStack_mature.fasta | wc -l
grep "^>" Pmea_ShortStack_mature.fasta | wc -l
```

## Merge the three mature miRNA FASTAs

```{bash}
cd ../data/10-shortRNA-ShortStack-comparison

cat Apul_ShortStack_mature.fasta Peve_ShortStack_mature.fasta Pmea_ShortStack_mature.fasta > merged_all_ShortStack_mature.fasta

head merged_all_ShortStack_mature.fasta
tail merged_all_ShortStack_mature.fasta
```

## BLASTs

### Make database for each species:

Apul

```{r, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../data/10-shortRNA-ShortStack-comparison/Apul_ShortStack_mature.fasta \
-dbtype nucl \
-out ../output/10-shortRNA-ShortStack-comparison/blasts/Apul-db/Apul_ShortStack_mature
```

Peve

```{r, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../data/10-shortRNA-ShortStack-comparison/Peve_ShortStack_mature.fasta \
-dbtype nucl \
-out ../output/10-shortRNA-ShortStack-comparison/blasts/Peve-db/Peve_ShortStack_mature
```

Pmea

```{r, engine='bash'}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../data/10-shortRNA-ShortStack-comparison/Pmea_ShortStack_mature.fasta \
-dbtype nucl \
-out ../output/10-shortRNA-ShortStack-comparison/blasts/Pmea-db/Pmea_ShortStack_mature
```

### Run Blastn

Apul to all
```{r, engine='bash'}

/home/shared/ncbi-blast-2.11.0+/bin/blastn \
-task blastn \
-query ../data/10-shortRNA-ShortStack-comparison/merged_all_ShortStack_mature.fasta \
-db ../output/10-shortRNA-ShortStack-comparison/blasts/Apul-db/Apul_ShortStack_mature \
-out ../output/10-shortRNA-ShortStack-comparison/Apul_to_all_blastn.tab \
-evalue 1E-2 \
-num_threads 40 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/10-shortRNA-ShortStack-comparison/Apul_to_all_blastn.tab
```

Peve to all
```{r, engine='bash'}

/home/shared/ncbi-blast-2.11.0+/bin/blastn \
-task blastn \
-query ../data/10-shortRNA-ShortStack-comparison/merged_all_ShortStack_mature.fasta \
-db ../output/10-shortRNA-ShortStack-comparison/blasts/Peve-db/Peve_ShortStack_mature \
-out ../output/10-shortRNA-ShortStack-comparison/Peve_to_all_blastn.tab \
-evalue 1E-2 \
-num_threads 40 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/10-shortRNA-ShortStack-comparison/Peve_to_all_blastn.tab
```

Pmea to all
```{r, engine='bash'}

/home/shared/ncbi-blast-2.11.0+/bin/blastn \
-task blastn \
-query ../data/10-shortRNA-ShortStack-comparison/merged_all_ShortStack_mature.fasta \
-db ../output/10-shortRNA-ShortStack-comparison/blasts/Pmea-db/Pmea_ShortStack_mature \
-out ../output/10-shortRNA-ShortStack-comparison/Pmea_to_all_blastn.tab \
-evalue 1E-2 \
-num_threads 40 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6

wc -l ../output/10-shortRNA-ShortStack-comparison/Pmea_to_all_blastn.tab
```

# Join BLAST tables

```{r}
apul_to_all_blastn <- read.table("../output/10-shortRNA-ShortStack-comparison/Apul_to_all_blastn.tab", sep="\t", header=FALSE)
peve_to_all_blastn <- read.table("../output/10-shortRNA-ShortStack-comparison/Peve_to_all_blastn.tab", sep="\t", header=FALSE)
pmea_to_all_blastn <- read.table("../output/10-shortRNA-ShortStack-comparison/Pmea_to_all_blastn.tab", sep="\t", header=FALSE)
```

Column labels: 
qseqid: Query sequence ID 
sseqid: Subject (database)
sequence ID pident: Percentage of identical matches 
length: Alignment length (number of base pairs or amino acids) 
mismatch: Number of mismatches 
gapopen: Number of gap openings 
qstart: Start of alignment in the query 
qend: End of alignment in the query 
sstart: Start of alignment in the subject 
send: End of alignment in the subject 
evalue: Expect value (number of hits expected by chance) 
bitscore: Bit score

```{r}
# Combine the three blast tables
combined_blastn <- rbind(apul_to_all_blastn, peve_to_all_blastn, pmea_to_all_blastn)

# Assign informative column labels
colnames(combined_blastn) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")

# Remove instances where a sequence matched to itself (e.g. from querying an Apul sequence against our combined database which contained all Apul sequences)
corrected_combined_blastn <- filter(combined_blastn, qseqid != sseqid)

# Remove any hits that have mismatches
filtered_combined_blastn <- filter(corrected_combined_blastn, mismatch==0)

# View
nrow(filtered_combined_blastn)
head(filtered_combined_blastn)

write.table(filtered_combined_blastn, "../output/10-shortRNA-ShortStack-comparison/filtered_combined_blast.tab", sep="\t", row.names=FALSE, quote=FALSE)
```

# Identify conserved miRNAs

Ok now we can start identifying conserved miRNAs. Keep in mind that this list of filtered, combined blastn hits contains duplicates because, for example, querying Apul sequences against a database containing Peve sequences is functionally the same as querying those Peve sequences against a databse which contains Apul. So, for example, this list would contain a hit matching Apul.seq1 to Peve.seq2, *and* a hit matching Peve.seq2 to Apul.seq1.


## Conserved across all three species (Apul, Peve, and Pmea)

First, lets find miRNAs conserved among all three species. These would show up as an miRNA from one species that has hits from both other species (e.g., Apul.seq1 has a hit from Peve *and* a hit from Pmea).
```{r}
# Find Apul miRNAs that have matches from both Peve and Pmea
present_in_all <- filtered_combined_blastn %>%
  # isolate Apul miRNAs with hits
  filter(!grepl("Porites_evermani|Pocillopora_meandrina", sseqid)) %>%
  group_by(sseqid) %>%
  filter(any(grepl("Porites_evermani", qseqid)) & any(grepl("Pocillopora_meandrina", qseqid)))

# View the miRNAs that match across all three species
# (recall this will include two entries for each conserved miRNA, it's Apul match in Peve, and its Apul match to Pmea)
head(present_in_all, nrow(present_in_all))
 
# Count the number of miRNAs conserved across all three species
paste("Number of miRNAs conserved across all three species:", nrow(distinct(present_in_all, sseqid)))
```

## Conserved among subsets of the three species

Now we want to find miRNAs that are conserved withing subsets of the three species

### Apul and Peve

Find Apul miRNAs that have hits to Peve miRNAs but *not* hits to Pmea miRNAs (that would make them conserved among all three species, which we've already identified)
```{r}
# Find Apul miRNAs that have matches from only Peve
present_in_apul_peve <- filtered_combined_blastn %>%
  # isolate Apul miRNAs with hits
  filter(!grepl("Porites_evermani|Pocillopora_meandrina", sseqid)) %>%
  group_by(sseqid) %>%
  # filter for hits to Peve only
  filter(any(grepl("Porites_evermani", qseqid)) & !any(grepl("Pocillopora_meandrina", qseqid)))

# View the miRNAs that match between Apul and Peve
head(present_in_apul_peve, nrow(present_in_apul_peve))
 
# Count the number of miRNAs conserved across the two species
paste("Number of miRNAs conserved in Apul and Peve:", nrow(distinct(present_in_apul_peve, sseqid)))
```


### Apul and Pmea

Find Apul miRNAs that have hits to Pmea miRNAs but *not* hits to Peve miRNAs

```{r}
# Find Apul miRNAs that have matches from only Pmea
present_in_apul_pmea <- filtered_combined_blastn %>%
  # isolate Apul miRNAs with hits
  filter(!grepl("Porites_evermani|Pocillopora_meandrina", sseqid)) %>%
  group_by(sseqid) %>%
  # filter for hits to Pmea only
  filter(!any(grepl("Porites_evermani", qseqid)) & any(grepl("Pocillopora_meandrina", qseqid)))

# View the miRNAs that match between Apul and Pmea
head(present_in_apul_pmea, nrow(present_in_apul_pmea))
 
# Count the number of miRNAs conserved across the two species
paste("Number of miRNAs conserved in Apul and Pmea:", nrow(distinct(present_in_apul_pmea, sseqid)))
```

### Peve and Pmea

Find Peve miRNAs that have hits to Pmea miRNAs but *not* hits to Apul miRNAs

```{r}
# Find Peve miRNAs that have matches from only Pmea
present_in_peve_pmea <- filtered_combined_blastn %>%
  # isolate Peve miRNAs with hits
  filter(grepl("Porites_evermani", sseqid)) %>%
  group_by(sseqid) %>%
  # filter for hits to Pmea only (note the Apul sequence IDs don't contain the species name, so we have to use a non-descriptive unique identifier for filtering)
  filter(!any(grepl("mature::N", qseqid)) & any(grepl("Pocillopora_meandrina", qseqid)))

# View the miRNAs that match between Peve and Pmea
head(present_in_peve_pmea, nrow(present_in_peve_pmea))
 
# Count the number of miRNAs conserved across the two species
paste("Number of miRNAs conserved in Peve and Pmea:", nrow(distinct(present_in_peve_pmea, sseqid)))
```


# Visualize

## Data munging of the results

```{r, engine='bash'}
cd ../data/10-shortRNA-ShortStack-comparison
grep "^>"  merged_all_ShortStack_mature.fasta | sed 's/^>//' > merged_all_ShortStack_mature_IDs.txt

head -5 merged_all_ShortStack_mature_IDs.txt
```
```{r}
# Read in and separate the ids of all miRNAs from the three species
merged_IDs <- readLines("../data/10-shortRNA-ShortStack-comparison/merged_all_ShortStack_mature_IDs.txt")

apul_IDs <- merged_IDs[grep("mature::N", merged_IDs)]
peve_IDs <- merged_IDs[grep("Porites_evermani", merged_IDs)]
pmea_IDs <- merged_IDs[grep("Pocillopora_meandrina", merged_IDs)]
length(apul_IDs)
length(peve_IDs)
length(pmea_IDs)
```

```{r}
# Assign shared miRNA IDs to conserved miRNAs

# Function to append IDs of matching miRNAs to the original query miRNA
append_IDs <- function(IDs_list, df) {
  appended_IDs_list <- vector("list", length(IDs_list))
  for (i in seq_along(IDs_list)) {
    matching_entries <- df$qseqid[df$sseqid == IDs_list[i]]
    if (length(matching_entries) > 0) {
      appended_IDs_list[[i]] <- paste(IDs_list[i], paste(matching_entries, collapse = "|"), sep = "|")
    } else {
      appended_IDs_list[[i]] <- IDs_list[i]
    }
  }
  return(appended_IDs_list)
}

# Apply the function to each set of conserved miRNAs
appendedIDs_apul_peve_pmea <- append_IDs(unique(present_in_all$sseqid), present_in_all)

appendedIDs_apul_peve <- append_IDs(unique(present_in_apul_peve$sseqid), present_in_apul_peve)
appendedIDs_apul_pmea <- append_IDs(unique(present_in_apul_pmea$sseqid), present_in_apul_pmea)
appendedIDs_peve_pmea <- append_IDs(unique(present_in_peve_pmea$sseqid), present_in_peve_pmea)

print(appendedIDs_apul_peve_pmea[1])
print(appendedIDs_apul_peve)
print(appendedIDs_apul_pmea)
print(appendedIDs_peve_pmea)

# combine the new appended IDs into a single list of conserved miRNAs
conserved_miRNAs_all_IDs <- c(appendedIDs_apul_peve_pmea, appendedIDs_apul_peve, appendedIDs_apul_pmea, appendedIDs_peve_pmea)
```

```{r}
# For each species list of miRNA IDs, replace species-specific IDs of conserved miRNAs with our newly generated appended IDs. This will created lists of miRNA IDs that have shared IDs for the conserved mRNAs
replace_entries <- function(spec_list, new_conserved_IDs) {
  # Iterate over each entry in spec_list
  for (i in seq_along(spec_list)) {
    # Check if the current entry in spec_list exists in any entry in new_conserved_IDs
    matching_entry <- new_conserved_IDs[grep(spec_list[i], new_conserved_IDs, fixed = TRUE)]
    # If a match is found, replace the entry in spec_list with the matching entry from new_conserved_IDs
    if (length(matching_entry) > 0) {
      spec_list[i] <- matching_entry[[1]]  # Replace with the first element of matching_entry
    }
  }
  return(spec_list)  # Return the modified spec_list
}

apul_mature_newconservedID <- replace_entries(apul_IDs, conserved_miRNAs_all_IDs)
peve_mature_newconservedID <- replace_entries(peve_IDs, conserved_miRNAs_all_IDs)
pmea_mature_newconservedID <- replace_entries(pmea_IDs, conserved_miRNAs_all_IDs)
```


## Venn diagram

```{r}
a <- list("A.pulchra" = apul_mature_newconservedID, 
          "P.evermanni" = peve_mature_newconservedID,
          "P.meandrina" = pmea_mature_newconservedID)

ggvenn(a)
```


