---
title: "Closest gene comparisons across species"
author: "Jill Ashey"
date: "2024-09-11"
output: html_document
---

This code will examine the overlap (if any) between the GO enrichment of the closest gene across species. See the following rmds for TopGO analysis by species: XXXXXXXX

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Biostrings) # Biostrings_2.62.0 
require(ggplot2)    # ggplot2_3.3.6
library(ggpubr)
library(ggseqlogo)  # install.packages("ggseqlogo")
library(reshape2)   # reshape2_1.4.4
library(data.table)
library(gghighlight) #install.packages("gghighlight")
library(plotrix)
library(dplyr)
library(tidyr)
library(tibble)
library(tidyverse)
library(ggvenn)
library(cowplot)
```

## piRNA

Compare terms identified in piRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_pi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_piRNA.csv")
apul_go_pi$Species <- "Acropora pulchra"
peve_go_pi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_piRNA.csv")
peve_go_pi$Species <- "Porites evermanni"
pmea_go_pi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_piRNA.csv")
pmea_go_pi$Species <- "Pocillopora tuahiniensis"

all_go_pi <- rbind(apul_go_pi, pmea_go_pi, peve_go_pi)
```

Subset by ontology 
```{r}
# Apul
apul_go_pi_bp <- apul_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_pi_cc <- apul_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_pi_mf <- apul_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_pi_bp <- peve_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_pi_cc <- peve_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_pi_mf <- peve_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_pi_bp <- pmea_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_pi_cc <- pmea_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_pi_mf <- pmea_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_pi <- list(APUL=apul_go_pi_bp$GeneOntologyIDs, PTUH=pmea_go_pi_bp$GeneOntologyIDs,PEVE=peve_go_pi_bp$GeneOntologyIDs)

bp_venn_pi <- ggvenn(GOs_BP_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_pi

# Cellular Components
GOs_CC_wide_pi <- list(APUL=apul_go_pi_cc$GeneOntologyIDs, PTUH=pmea_go_pi_cc$GeneOntologyIDs,PEVE=peve_go_pi_cc$GeneOntologyIDs)

cc_venn_pi <- ggvenn(GOs_CC_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_pi

# Molecular Functions
GOs_MF_wide_pi <- list(APUL=apul_go_pi_mf$GeneOntologyIDs, PTUH=pmea_go_pi_mf$GeneOntologyIDs,PEVE=peve_go_pi_mf$GeneOntologyIDs)

mf_venn_pi <- ggvenn(GOs_MF_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_pi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_pi <- all_go_pi %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_pi_df <- all_go_pi %>%
  filter(GeneOntologyIDs %in% shared_go_pi$GeneOntologyIDs)

# Plot 
GO_plot_pi <- ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_pi

ggsave(filename = "../figures/piRNA_shared_GOs_plot.png", GO_plot_pi, width = 25, height = 25)
ggsave(filename = "../figures/piRNA_shared_GOs_plot.pdf", GO_plot_pi, width = 25, height = 25)
```

I like this plot a lot!

## miRNA

Compare terms identified in miRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_mi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_miRNA.csv")
apul_go_mi$Species <- "Acropora pulchra"
peve_go_mi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_miRNA.csv")
peve_go_mi$Species <- "Porites evermanni"
pmea_go_mi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_miRNA.csv")
pmea_go_mi$Species <- "Pocillopora tuahiniensis"

all_go_mi <- rbind(apul_go_mi, pmea_go_mi, peve_go_mi)
```

Subset by ontology 
```{r}
# Apul
apul_go_mi_bp <- apul_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_mi_cc <- apul_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_mi_mf <- apul_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_mi_bp <- peve_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_mi_cc <- peve_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_mi_mf <- peve_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_mi_bp <- pmea_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_mi_cc <- pmea_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_mi_mf <- pmea_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_mi <- list(APUL=apul_go_mi_bp$GeneOntologyIDs, PTUH=pmea_go_mi_bp$GeneOntologyIDs,PEVE=peve_go_mi_bp$GeneOntologyIDs)

bp_venn_mi <- ggvenn(GOs_BP_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_mi

# Cellular Components
GOs_CC_wide_mi <- list(APUL=apul_go_mi_cc$GeneOntologyIDs, PTUH=pmea_go_mi_cc$GeneOntologyIDs,PEVE=peve_go_mi_cc$GeneOntologyIDs)

cc_venn_mi <- ggvenn(GOs_CC_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_mi

# Molecular Functions
GOs_MF_wide_mi <- list(APUL=apul_go_mi_mf$GeneOntologyIDs, PTUH=pmea_go_mi_mf$GeneOntologyIDs,PEVE=peve_go_mi_mf$GeneOntologyIDs)

mf_venn_mi <- ggvenn(GOs_MF_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_mi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_mi <- all_go_mi %>%
  group_by(GeneOntologyIDs) %>%  # Group by both GeneOntologyIDs and ontology
  summarise(unique_species = n_distinct(Species), .groups = 'keep') %>%  # Count distinct species for each GO term and ontology
  filter(unique_species > 1)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_mi_df <- all_go_mi %>%
  filter(GeneOntologyIDs %in% shared_go_mi$GeneOntologyIDs)

# Identify terms associated with both CC and MF, and only keep CC
shared_go_mi_df <- shared_go_mi_df %>%
  group_by(GeneOntologyIDs) %>%
  mutate(cc_present = any(ontology == "Cellular Components")) %>%
  filter(!(ontology == "Molecular Functions" & cc_present)) %>%
  ungroup() %>%
  dplyr::select(-cc_present)

# Plot 
GO_plot_mi <- ggplot(shared_go_mi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_mi

ggsave(filename = "../figures/miRNA_shared_GOs_plot.png", GO_plot_mi, width = 25, height = 30)
ggsave(filename = "../figures/miRNA_shared_GOs_plot.pdf", GO_plot_mi, width = 25, height = 30)
```

## lncRNAs 

Compare terms identified in piRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_lnc <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_lncRNA.csv")
apul_go_lnc$Species <- "Acropora pulchra"
peve_go_lnc <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_lncRNA.csv")
peve_go_lnc$Species <- "Porites evermanni"
pmea_go_lnc <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_lncRNA.csv")
pmea_go_lnc$Species <- "Pocillopora tuahiniensis"

all_go_lnc <- rbind(apul_go_lnc, pmea_go_lnc, peve_go_lnc)
```

Subset by ontology 
```{r}
# Apul
apul_go_lnc_bp <- apul_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_lnc_cc <- apul_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_lnc_mf <- apul_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_lnc_bp <- peve_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_lnc_cc <- peve_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_lnc_mf <- peve_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_lnc_bp <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_lnc_cc <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_lnc_mf <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_lnc <- list(APUL=apul_go_lnc_bp$GeneOntologyIDs, PTUH=pmea_go_lnc_bp$GeneOntologyIDs,PEVE=peve_go_lnc_bp$GeneOntologyIDs)

bp_venn_lnc <- ggvenn(GOs_BP_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_lnc

# Cellular Components
GOs_CC_wide_lnc <- list(APUL=apul_go_lnc_cc$GeneOntologyIDs, PTUH=pmea_go_lnc_cc$GeneOntologyIDs,PEVE=peve_go_lnc_cc$GeneOntologyIDs)

cc_venn_lnc <- ggvenn(GOs_CC_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_lnc

# Molecular Functions
GOs_MF_wide_lnc <- list(APUL=apul_go_lnc_mf$GeneOntologyIDs, PTUH=pmea_go_lnc_mf$GeneOntologyIDs,PEVE=peve_go_lnc_mf$GeneOntologyIDs)

mf_venn_lnc <- ggvenn(GOs_MF_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_lnc
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_lnc <- all_go_lnc %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_lnc_df <- all_go_lnc %>%
  filter(GeneOntologyIDs %in% shared_go_lnc$GeneOntologyIDs)

# Plot 
GO_plot_lnc <- ggplot(shared_go_lnc_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
    panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_lnc

ggsave(filename = "../figures/lncRNA_shared_GOs_plot.png", GO_plot_lnc, width = 35, height = 49)
ggsave(filename = "../figures/lncRNA_shared_GOs_plot.pdf", GO_plot_lnc, width = 35, height = 49)
```

## All ncRNAs 

Filter out extra information from `shared_go_pi_df`, `shared_go_mi_df`, and `shared_go_lnc_df`
```{r}
# Look at dimensions and column names for each df 
dim(shared_go_pi_df)
dim(shared_go_mi_df)
dim(shared_go_lnc_df)
colnames(shared_go_pi_df)
colnames(shared_go_mi_df)
colnames(shared_go_lnc_df)

# Select certain columns 
filt_shared_go_pi_df <- shared_go_pi_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, pi_chrom, pi_start, pi_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = pi_chrom, ncrna_start = pi_start, ncrna_end = pi_end) %>%
  mutate(ncrna = "piRNA")

filt_shared_go_mi_df <- shared_go_mi_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, m_chrom, m_start, m_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = m_chrom, ncrna_start = m_start, ncrna_end = m_end) %>%
  mutate(ncrna = "miRNA")

filt_shared_go_lnc_df <- shared_go_lnc_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, l_chrom, l_start, l_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = l_chrom, ncrna_start = l_start, ncrna_end = l_end) %>%
  mutate(ncrna = "lncRNA")

# Bind dfs together 
all_shared <- rbind(filt_shared_go_pi_df, filt_shared_go_mi_df, filt_shared_go_lnc_df)

# Save as csv 
write.csv(all_shared, "../output/19-bedtools-closest-gene/ncRNAs_shared_go_by_species.csv")
```

Plot 
```{r}
GO_plot_all <- ggplot(all_shared, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_all

ggsave(filename = "../figures/all_ncRNAs_shared_GOs_plot.png", GO_plot_all, width = 35, height = 49)
ggsave(filename = "../figures/all_ncRNAs_shared_GOs_plot.pdf", GO_plot_all, width = 35, height = 49)
```

Plot only miRNAs and piRNAs
```{r}
small_shared <- rbind(filt_shared_go_pi_df, filt_shared_go_mi_df)

GO_plot_smol <- ggplot(small_shared, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_smol

ggsave(filename = "../figures/small_ncRNAs_shared_GOs_plot.png", GO_plot_smol, width = 25, height = 35)
ggsave(filename = "../figures/small_ncRNAs_shared_GOs_plot.pdf", GO_plot_smol, width = 25, height = 35)
```


#### Plot Biological Processes only 
Plot 
```{r}
all_shared_BP <- all_shared %>%
  dplyr::filter(ontology == "Biological Processes")

GO_plot_all_BP <- ggplot(all_shared_BP, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_all_BP

ggsave(filename = "../figures/all_ncRNAs_shared_GO_BP_plot.png", GO_plot_all_BP, width = 35, height = 49)
ggsave(filename = "../figures/all_ncRNAs_shared_GOs_BP_plot.pdf", GO_plot_all_BP, width = 35, height = 49)
```

Plot only BP GO terms for miRNAs and piRNAs
```{r}
small_shared <- rbind(filt_shared_go_pi_df, filt_shared_go_mi_df)

small_shared_BP <- small_shared %>%
  dplyr::filter(ontology == "Biological Processes")

GO_plot_smol_BP <- ggplot(small_shared_BP, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 22, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_smol_BP

ggsave(filename = "../figures/small_ncRNAs_shared_GO_BP_plot.png", GO_plot_smol_BP, width = 35, height = 25)
ggsave(filename = "../figures/small_ncRNAs_shared_GO_BP_plot.pdf", GO_plot_smol_BP, width = 35, height = 25)
```

Plot only BP GO terms for lncRNAs
```{r}
shared_go_lnc_df_BP <- all_shared %>%
  dplyr::filter(ontology == "Biological Processes") %>%
  dplyr::filter(ncrna == "lncRNA")

GO_plot_lnc_BP <- ggplot(shared_go_lnc_df_BP, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 22, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_lnc_BP

ggsave(filename = "../figures/lncRNAs_shared_GO_BP_plot.png", GO_plot_lnc_BP, width = 35, height = 25)
ggsave(filename = "../figures/lncRNAs_shared_GO_BP_plot.pdf", GO_plot_lnc_BP, width = 35, height = 25)
```





## Plot lengths for each species 

I am going to make length plots for each species for lncRNAs, miRNAs and piRNAs so that the format is the same for everything. Also going to make venn diagrams for shared sequences for lncRNAs and miRNAs here so that the format is the same and I can join them in one figure. 


### miRNAs 

Kathleen wrote code for the miRNA [length comparison](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/10-shortRNA-ShortStack-comparison.Rmd). I'm going to modify her code 

Read in length data for miRNAs
```{r}
Apul_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Apul_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
Peve_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Peve_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
Pmea_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Pmea_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
```

Combine dfs
```{r}
# Add a new column to each data frame to label the source file
Apul_lengths_mi <- Apul_lengths_mi %>% mutate(Species = 'A. pulchra')
Peve_lengths_mi <- Peve_lengths_mi %>% mutate(Species = 'P. evermanni')
Pmea_lengths_mi <- Pmea_lengths_mi %>% mutate(Species = 'P. tuahiniensis')

# Combine the data frames into one
all_lengths_mi <- rbind(Apul_lengths_mi, Peve_lengths_mi, Pmea_lengths_mi)
```

Count miRNA lengths by species
```{r}
count_by_group_mi <- all_lengths_mi %>%
  dplyr::count(Species, length)
count_by_group_mi
```

Plot miRNA lengths of all species
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_mi <- ggplot(count_by_group, aes(x = length, y = n, fill = Species)) + 
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  scale_y_continuous(expand = c(0, 0)) +   # Bars sit on x-axis
  labs(x = "Length (bp)", y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.position = "none"                                # Remove legend
  ); length_plot_mi


ggsave(filename = "../figures/miRNA_lengths.png", length_plot_mi, width = 34, height = 20)
ggsave(filename = "../figures/miRNA_lengths.pdf", length_plot_mi, width = 34, height = 20)
```

The actual Venn diagram was made by Kathleen [here](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/10-shortRNA-ShortStack-comparison.Rmd). She did some complex code to get the venn to work so I am just doing to pull the numbers from the venn to make the plot format uniform.

Make lists for venn
```{r}
# Define elements for each set based on data, correctly reflecting overlaps
set_APUL <- c(paste0("APUL", 1:32), "APUL_PEVE", "APUL_PMEA", paste0("APUL_PEVE_PMEA", 1:4))
set_PMEA <- c(paste0("PMEA", 1:31), "APUL_PMEA", "PMEA_PEVE", paste0("APUL_PEVE_PMEA", 1:4))
set_PEVE <- c(paste0("PEVE", 1:40), "APUL_PEVE", "PMEA_PEVE", paste0("APUL_PEVE_PMEA", 1:4))

# Combine the sets into a named list
venn_sets <- list(
  "A. pulchra" = set_APUL,
  "P. tuahiniensis" = set_PMEA,
  "P. evermanni" = set_PEVE
)
```

Plot overlap of miRNAs across spp
```{r}
venn_mi <- ggvenn(
  venn_sets, 
  fill_color = species_colors_nolabel,  # Customize the fill colors
  stroke_size = 1.5, # Outline size of the circles
  set_name_size = F,         # Size of the set labels
  show_percentage = F,
  stroke_alpha = 2,
  text_size = 20
); venn_mi

ggsave(filename = "../figures/miRNA_overlap_venn.png", venn_mi, width = 34, height = 20)
ggsave(filename = "../figures/miRNA_overlap_venn.pdf", venn_mi, width = 34, height = 20)
```

### piRNAs

Javi wrote code for the piRNA [length comparison](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/12-piRNA-comparizon.Rmd). I'm going to modify his code.

Read in length data for piRNAs 
```{r}
APUL_L = list.files(path = "../../D-Apul/output/18-Apul-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

PMEA_L = list.files(path = "../../F-Pmea/output/18-Pmea-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

PEVE_L = list.files(path = "../../E-Peve/output/18-Peve-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

L = c(APUL_L,PMEA_L,PEVE_L)
names <- L %>%
  substring(.,75) %>%
  gsub("_piRNA_length.txt", "", .)

DFs = lapply(L, function(x) {
  data <- read.table(paste0(x), comment.char = "")
  colnames(data) <- c("freq", "length")
  return(data)
})
```

Count piRNA lengths by species  
```{r}
combo = Reduce(function(...) merge(..., by = "length", all=T), DFs)
colnames(combo) <- c("length", names)
combo <- combo %>%
  mutate(ACR = rowSums(combo[,2:6])) %>%
  mutate(POC = rowSums(combo[,7:11])) %>%
  mutate(POR = rowSums(combo[,12:14])) %>%
  select(length,ACR,POC,POR) %>%
  pivot_longer(c("ACR", "POC", "POR"), values_to = "sums", names_to = "group") %>%
  mutate(Species = case_when(
    group == "ACR" ~ "A. pulchra", 
    group == "POR" ~ "P. evermanni",
    group == "POC" ~ "P. tuahiniensis"))
```

Plot piRNA lengths of all species
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_pi <- ggplot(combo, aes(x = length, y = sums, fill = Species)) + 
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  scale_y_continuous(expand = c(0, 0)) +   # Bars sit on x-axis
  scale_x_continuous(breaks = unique(combo$length)) +  # Ensure every "length" value appears on the x-axis
  labs(x = "Length (bp)", y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.title = element_text(size = 40, face = "bold"),  # Increase legend title size
    legend.text = element_text(size = 40),
    legend.position = "none"                              
  ) +
  guides(fill = guide_legend(override.aes = list(size = 20)));  # Increase the size of legend squares
length_plot_pi

ggsave(filename = "../figures/piRNA_lengths.png", length_plot_pi, width = 34, height = 20)
ggsave(filename = "../figures/piRNA_lengths.pdf", length_plot_pi, width = 34, height = 20)
```

### lncRNAs 

Calculate length from bed files
```{r}
apul_lnc_bed <- read.delim("../../D-Apul/output/05.33-lncRNA-discovery/Apul_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Acropora pulchra")

peve_lnc_bed <- read.delim("../../E-Peve/output/Peve_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Porites evermanni")

pmea_lnc_bed <- read.delim("../../F-Pmea/output/02-lncRNA-discovery/Pmea_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Pocillopora tuahiniensis")

# Combine the data frames into one
all_lengths_lnc <- rbind(apul_lnc_bed, peve_lnc_bed, pmea_lnc_bed)
```

Define custom bins so that the lncRNAs are grouped into 200-300 bp, 300-400 bp, etc
```{r}
all_lengths_lnc <- all_lengths_lnc %>%
  mutate(length_bin = cut(
    length,
    breaks = c(200, 300, 400, 500, 600, 700, 800, 900, 1000, Inf),  # Define custom breaks
    labels = c("200-300", "300-400", "400-500", "500-600", "600-700", "700-800", "800-900", "900-1000", ">1000"),  # Custom labels for each bin
    include.lowest = TRUE,  # Include the lowest value in the first interval
    right = FALSE           # Make the intervals left-closed (e.g., 200-300 includes 200, but not 300)
  ))
```

Calculate frequency of lncRNAs in each bin by species
```{r}
length_freq <- all_lengths_lnc %>%
  group_by(Species, length_bin) %>%
  summarize(frequency = n(), .groups = 'drop')
```

Plot lncRNA lengths for all species 
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_lnc <- ggplot(length_freq, aes(x = length_bin, y = frequency, fill = Species)) +
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  # Use scale_x_discrete instead of scale_x_continuous
  scale_x_discrete(name = "Length (bp)") +  # Ensure every "length_bin" value appears on the x-axis
    scale_y_continuous(name = "Read count", expand = c(0, 0)) +  # Remove padding on y-axis so bars sit on x-axis

  labs(y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis text for better readability
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.title = element_text(size = 40, face = "bold"),  # Increase legend title size
    legend.text = element_text(size = 40),
    legend.position = "none"                              
  ) +
  guides(fill = guide_legend(override.aes = list(size = 20)));  # Increase the size of legend squares

length_plot_lnc

ggsave(filename = "../figures/lncRNA_lengths.png", length_plot_lnc, width = 34, height = 20)
ggsave(filename = "../figures/lncRNA_lengths.pdf", length_plot_lnc, width = 34, height = 20)
```


### ALL ncRNAs 

Still need lncRNA length and similarity data. Once I have that, I will put the length and venn plots in one figure here. 

Plot RNA lengths + venns
```{r}
plot <- plot_grid(length_plot_lnc, venn_mi, length_plot_mi, venn_mi, length_plot_pi, labels = c("A", "B", "C", "D", "E"), label_size = 80, nrow = 3, ncol = 2)

ggsave(filename = "../figures/ncRNA_lengths_overlap.png", plot, width = 55, height = 45, limitsize = F)
ggsave(filename = "../figures/ncRNA_lengths_overlap.pdf", plot, width = 55, height = 45, limitsize = F)
```






