---
title: "Closest gene comparisons across species"
author: "Jill Ashey"
date: "2024-09-11"
output: html_document
---

This code will examine the overlap (if any) between the GO enrichment of the closest gene across species. See the following rmds for TopGO analysis by species: XXXXXXXX

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Biostrings) # Biostrings_2.62.0 
require(ggplot2)    # ggplot2_3.3.6
library(ggpubr)
library(ggseqlogo)  # install.packages("ggseqlogo")
library(reshape2)   # reshape2_1.4.4
library(data.table)
library(gghighlight) #install.packages("gghighlight")
library(plotrix)
library(dplyr)
library(tidyr)
library(tibble)
library(tidyverse)
library(ggvenn)
library(cowplot)
library(gridExtra)
library(ComplexUpset)
```

## piRNA

Compare terms identified in piRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_pi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_piRNA.csv")
apul_go_pi$Species <- "Acropora pulchra"
peve_go_pi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_piRNA.csv")
peve_go_pi$Species <- "Porites evermanni"
pmea_go_pi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_piRNA.csv")
pmea_go_pi$Species <- "Pocillopora tuahiniensis"

all_go_pi <- rbind(apul_go_pi, pmea_go_pi, peve_go_pi)
```

Subset by ontology 
```{r}
# Apul
apul_go_pi_bp <- apul_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_pi_cc <- apul_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_pi_mf <- apul_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_pi_bp <- peve_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_pi_cc <- peve_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_pi_mf <- peve_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_pi_bp <- pmea_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_pi_cc <- pmea_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_pi_mf <- pmea_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_pi <- list(APUL=apul_go_pi_bp$GeneOntologyIDs, PTUH=pmea_go_pi_bp$GeneOntologyIDs,PEVE=peve_go_pi_bp$GeneOntologyIDs)

bp_venn_pi <- ggvenn(GOs_BP_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_pi

# Cellular Components
GOs_CC_wide_pi <- list(APUL=apul_go_pi_cc$GeneOntologyIDs, PTUH=pmea_go_pi_cc$GeneOntologyIDs,PEVE=peve_go_pi_cc$GeneOntologyIDs)

cc_venn_pi <- ggvenn(GOs_CC_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_pi

# Molecular Functions
GOs_MF_wide_pi <- list(APUL=apul_go_pi_mf$GeneOntologyIDs, PTUH=pmea_go_pi_mf$GeneOntologyIDs,PEVE=peve_go_pi_mf$GeneOntologyIDs)

mf_venn_pi <- ggvenn(GOs_MF_wide_pi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_pi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_pi <- all_go_pi %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_pi_df <- all_go_pi %>%
  filter(GeneOntologyIDs %in% shared_go_pi$GeneOntologyIDs)

# Plot 
GO_plot_pi <- ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_pi

ggsave(filename = "../figures/piRNA_shared_GOs_plot.png", GO_plot_pi, width = 25, height = 25)
ggsave(filename = "../figures/piRNA_shared_GOs_plot.pdf", GO_plot_pi, width = 25, height = 25)
```

I like this plot a lot!

Plot only BP 
```{r}
shared_go_pi_df_bp <- shared_go_pi_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_BP_plot_pi <- ggplot(shared_go_pi_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 30, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(angle = 75, hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_BP_plot_pi

ggsave(filename = "../figures/piRNA_shared_GO_BP_plot.png", GO_BP_plot_pi, width = 25, height = 35)
ggsave(filename = "../figures/piRNA_shared_GO_BP_plot.pdf", GO_BP_plot_pi, width = 25, height = 35)
```


## miRNA

Compare terms identified in miRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_mi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_miRNA.csv")
apul_go_mi$Species <- "Acropora pulchra"
peve_go_mi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_miRNA.csv")
peve_go_mi$Species <- "Porites evermanni"
pmea_go_mi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_miRNA.csv")
pmea_go_mi$Species <- "Pocillopora tuahiniensis"

all_go_mi <- rbind(apul_go_mi, pmea_go_mi, peve_go_mi)
```

Subset by ontology 
```{r}
# Apul
apul_go_mi_bp <- apul_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_mi_cc <- apul_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_mi_mf <- apul_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_mi_bp <- peve_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_mi_cc <- peve_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_mi_mf <- peve_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_mi_bp <- pmea_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_mi_cc <- pmea_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_mi_mf <- pmea_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_mi <- list(APUL=apul_go_mi_bp$GeneOntologyIDs, PTUH=pmea_go_mi_bp$GeneOntologyIDs,PEVE=peve_go_mi_bp$GeneOntologyIDs)

bp_venn_mi <- ggvenn(GOs_BP_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_mi

# Cellular Components
GOs_CC_wide_mi <- list(APUL=apul_go_mi_cc$GeneOntologyIDs, PTUH=pmea_go_mi_cc$GeneOntologyIDs,PEVE=peve_go_mi_cc$GeneOntologyIDs)

cc_venn_mi <- ggvenn(GOs_CC_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_mi

# Molecular Functions
GOs_MF_wide_mi <- list(APUL=apul_go_mi_mf$GeneOntologyIDs, PTUH=pmea_go_mi_mf$GeneOntologyIDs,PEVE=peve_go_mi_mf$GeneOntologyIDs)

mf_venn_mi <- ggvenn(GOs_MF_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_mi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_mi <- all_go_mi %>%
  group_by(GeneOntologyIDs) %>%  # Group by both GeneOntologyIDs and ontology
  summarise(unique_species = n_distinct(Species), .groups = 'keep') %>%  # Count distinct species for each GO term and ontology
  filter(unique_species > 1)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_mi_df <- all_go_mi %>%
  filter(GeneOntologyIDs %in% shared_go_mi$GeneOntologyIDs)

# Identify terms associated with both CC and MF, and only keep CC
shared_go_mi_df <- shared_go_mi_df %>%
  group_by(GeneOntologyIDs) %>%
  mutate(cc_present = any(ontology == "Cellular Components")) %>%
  filter(!(ontology == "Molecular Functions" & cc_present)) %>%
  ungroup() %>%
  dplyr::select(-cc_present)

# Plot 
GO_plot_mi <- ggplot(shared_go_mi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black", angle = 75),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ); GO_plot_mi

ggsave(filename = "../figures/miRNA_shared_GOs_plot.png", GO_plot_mi, width = 25, height = 30)
ggsave(filename = "../figures/miRNA_shared_GOs_plot.pdf", GO_plot_mi, width = 25, height = 30)
```

Plot only BP 
```{r}
shared_go_mi_df_bp <- shared_go_mi_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_BP_plot_mi <- ggplot(shared_go_mi_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 30, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(angle = 75, hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_BP_plot_mi

ggsave(filename = "../figures/miRNA_shared_GO_BP_plot.png", GO_BP_plot_mi, width = 25, height = 35)
ggsave(filename = "../figures/miRNA_shared_GO_BP_plot.pdf", GO_BP_plot_mi, width = 25, height = 35)
```

## lncRNAs 

Compare terms identified in piRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_lnc <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_lncRNA.csv")
apul_go_lnc$Species <- "Acropora pulchra"
peve_go_lnc <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_lncRNA.csv")
peve_go_lnc$Species <- "Porites evermanni"
pmea_go_lnc <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_lncRNA.csv")
pmea_go_lnc$Species <- "Pocillopora tuahiniensis"

all_go_lnc <- rbind(apul_go_lnc, pmea_go_lnc, peve_go_lnc)
```

Subset by ontology 
```{r}
# Apul
apul_go_lnc_bp <- apul_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_lnc_cc <- apul_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_lnc_mf <- apul_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_lnc_bp <- peve_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_lnc_cc <- peve_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_lnc_mf <- peve_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_lnc_bp <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_lnc_cc <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_lnc_mf <- pmea_go_lnc %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_lnc <- list(APUL=apul_go_lnc_bp$GeneOntologyIDs, PTUH=pmea_go_lnc_bp$GeneOntologyIDs,PEVE=peve_go_lnc_bp$GeneOntologyIDs)

bp_venn_lnc <- ggvenn(GOs_BP_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_lnc

# Cellular Components
GOs_CC_wide_lnc <- list(APUL=apul_go_lnc_cc$GeneOntologyIDs, PTUH=pmea_go_lnc_cc$GeneOntologyIDs,PEVE=peve_go_lnc_cc$GeneOntologyIDs)

cc_venn_lnc <- ggvenn(GOs_CC_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_lnc

# Molecular Functions
GOs_MF_wide_lnc <- list(APUL=apul_go_lnc_mf$GeneOntologyIDs, PTUH=pmea_go_lnc_mf$GeneOntologyIDs,PEVE=peve_go_lnc_mf$GeneOntologyIDs)

mf_venn_lnc <- ggvenn(GOs_MF_wide_lnc, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_lnc
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_lnc <- all_go_lnc %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_lnc_df <- all_go_lnc %>%
  filter(GeneOntologyIDs %in% shared_go_lnc$GeneOntologyIDs)

# Plot 
GO_plot_lnc <- ggplot(shared_go_lnc_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
    panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_lnc

ggsave(filename = "../figures/lncRNA_shared_GOs_plot.png", GO_plot_lnc, width = 35, height = 49)
ggsave(filename = "../figures/lncRNA_shared_GOs_plot.pdf", GO_plot_lnc, width = 35, height = 49)
```

Plot only BP
```{r}
shared_go_lnc_df_bp <- shared_go_lnc_df %>%
  dplyr::filter(ontology == "Biological Processes")
 
# Plot 
GO_BP_plot_lnc <- ggplot(shared_go_lnc_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 30, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(angle = 75, hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 200)        # Add space below the plot
  ); GO_BP_plot_lnc

ggsave(filename = "../figures/lncRNA_shared_GO_BP_plot.png", GO_BP_plot_lnc, width = 45, height = 35)
ggsave(filename = "../figures/lncRNA_shared_GO_BP_plot.pdf", GO_BP_plot_lnc, width = 45, height = 35)
```

## All ncRNAs 

Filter out extra information from `shared_go_pi_df`, `shared_go_mi_df`, and `shared_go_lnc_df`
```{r}
# Look at dimensions and column names for each df 
dim(shared_go_pi_df)
dim(shared_go_mi_df)
dim(shared_go_lnc_df)
colnames(shared_go_pi_df)
colnames(shared_go_mi_df)
colnames(shared_go_lnc_df)

# Select certain columns 
filt_shared_go_pi_df <- shared_go_pi_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, pi_chrom, pi_start, pi_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = pi_chrom, ncrna_start = pi_start, ncrna_end = pi_end) %>%
  mutate(ncrna = "piRNA")

filt_shared_go_mi_df <- shared_go_mi_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, m_chrom, m_start, m_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = m_chrom, ncrna_start = m_start, ncrna_end = m_end) %>%
  mutate(ncrna = "miRNA")

filt_shared_go_lnc_df <- shared_go_lnc_df %>%
  dplyr::select(c(gene_name, GeneOntologyIDs, Term, Annotated, Significant, Expected, Fisher, ontology, l_chrom, l_start, l_end, gf_chrom, gf_type, gf_start, gf_end, overlap, prop.sig.genes, Species)) %>%
  rename(ncrna_chrom = l_chrom, ncrna_start = l_start, ncrna_end = l_end) %>%
  mutate(ncrna = "lncRNA")

# Bind dfs together 
all_shared <- rbind(filt_shared_go_pi_df, filt_shared_go_mi_df, filt_shared_go_lnc_df)

# Save as csv 
write.csv(all_shared, "../output/19-bedtools-closest-gene/ncRNAs_shared_go_by_species.csv")
```

Plot 
```{r}
GO_plot_all <- ggplot(all_shared, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 17, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), vars(ncrna), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"
  ) +
  coord_flip(); GO_plot_all

ggsave(filename = "../figures/all_ncRNAs_shared_GOs_plot.png", GO_plot_all, width = 35, height = 49)
ggsave(filename = "../figures/all_ncRNAs_shared_GOs_plot.pdf", GO_plot_all, width = 35, height = 49)
```

## Plot lengths for each species 

I am going to make length plots for each species for lncRNAs, miRNAs and piRNAs so that the format is the same for everything. Also going to make heatmap-esque diagrams for shared sequences for lncRNAs and miRNAs here so that the format is the same and I can join them in one figure. 


### miRNAs 

Kathleen wrote code for the miRNA [length comparison](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/10-shortRNA-ShortStack-comparison.Rmd). I'm going to modify her code 

Read in length data for miRNAs
```{r}
Apul_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Apul_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
Peve_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Peve_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
Pmea_lengths_mi <- read.table("../data/10-shortRNA-ShortStack-comparison/Pmea_ShortStack_mature_lengths.txt", header = FALSE, col.names = "length")
```

Combine dfs
```{r}
# Add a new column to each data frame to label the source file
Apul_lengths_mi <- Apul_lengths_mi %>% mutate(Species = 'A. pulchra')
Peve_lengths_mi <- Peve_lengths_mi %>% mutate(Species = 'P. evermanni')
Pmea_lengths_mi <- Pmea_lengths_mi %>% mutate(Species = 'P. tuahiniensis')

# Combine the data frames into one
all_lengths_mi <- rbind(Apul_lengths_mi, Peve_lengths_mi, Pmea_lengths_mi)
```

Count miRNA lengths by species
```{r}
count_by_group_mi <- all_lengths_mi %>%
  dplyr::count(Species, length)
count_by_group_mi
```

Plot miRNA lengths of all species
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_mi <- ggplot(count_by_group_mi, aes(x = length, y = n, fill = Species)) + 
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  scale_y_continuous(expand = c(0, 0)) +   # Bars sit on x-axis
  labs(x = "Length (bp)", y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.position = "none"                                # Remove legend
  ); length_plot_mi


ggsave(filename = "../figures/miRNA_lengths.png", length_plot_mi, width = 34, height = 20)
ggsave(filename = "../figures/miRNA_lengths.pdf", length_plot_mi, width = 34, height = 20)
```

The actual Venn diagram was made by Kathleen [here](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/10-shortRNA-ShortStack-comparison.Rmd). She did some complex code to get the venn to work so I am just doing to pull the numbers from the venn and make a new plot. We discussed making a heatmap looking plot instead of a venn diagram. I probably won't incorporate colors into the heatmap (for now), as the values cannot be proportions because each species has a different number of miRNAs and thus will have a different proportion shared between species. For example, Apul has 38 miRNAs and Peve has 37. They share 5 miRNAs between them; for Apul, the proportion would be 5/38 = 0.13 and for Peve, the proportion would be 0.14. In the boxes, I am going to have the number of miRNAs shared between that set of species. 
```{r}
# Create the data frame for the 3x3 grid
data <- data.frame(
  x = c("A. pulchra", "P. tuahiniensis", "P. evermanni", "A. pulchra", "P. tuahiniensis", 
        "P. evermanni", "A. pulchra", "P. tuahiniensis", "P. evermanni"),
  y = c("A. pulchra", "A. pulchra", "A. pulchra", "P. tuahiniensis", "P. tuahiniensis", 
        "P. tuahiniensis", "P. evermanni", "P. evermanni", "P. evermanni"),
  value = c(38, 5, 5, 5, 46, 5, 5, 5, 37)
)

# Add a column to identify repetitive (lower triangle) comparisons
data <- data %>%
  mutate(
    repetitive = ifelse(x > y, "yes", "no"),    # Flag repetitive squares
    shared = ifelse(repetitive == "yes", NA, value)  # Assign NA to repetitive cells for color gradient
  )

# Convert x and y to factors to control order
#data$x <- factor(data$x, levels = c("A. pulchra", "P. tuahiniensis", "P. evermanni"))
#data$y <- factor(data$y, levels = c("A. pulchra", "P. tuahiniensis", "P. evermanni"))

tile_overlap_mi <- ggplot(data, aes(x = x, y = y)) +
  geom_tile(aes(fill = repetitive), color = "black", size = 1) +               # Color tiles based on repetitive status
  geom_tile(data = subset(data, repetitive == "yes"),                # Overlay grey repetitive tiles
            fill = "gray", color = "black", size = 1) +
  geom_text(aes(label = ifelse(repetitive == "no", value, "")),      # Display values only for non-repetitive squares
            size = 25) +                                             # Add text in the middle of each tile
  #scale_fill_gradient(low = "blue", high = "red", na.value = "gray80") +  # Gradient for values, grey for NA
  scale_fill_manual(values = c("yes" = "gray80", "no" = "white"),    # Grey out repetitive squares, white for others
                    name = "Repetitive") +   
  theme_minimal() +
  theme(
    axis.title = element_blank(),                                    # Remove axis titles
    axis.text = element_text(size = 50, colour = "black", face = "italic"),              # Customize axis text
    axis.text.x = element_text(angle = 75, hjust = 1),  # Rotate and adjust label position
    #axis.line = element_line(size = 5, color = "black"),
    #panel.grid = element_blank(),                                    # Remove grid lines
    #panel.border = element_blank(),                                   # Remove panel borders
    legend.position = "none") +       
  coord_fixed(); tile_overlap_mi

ggsave(filename = "../figures/miRNA_tile_overlap.png", tile_overlap_mi, width = 25, height = 35)
ggsave(filename = "../figures/miRNA_tile_overlap.pdf", tile_overlap_mi, width = 25, height = 35)
```

Not a huge fan of the tile plot so going to make an upset plot as well. I need a list of all the miRNAs and then either a 1 if present or a 0 if absent across species. First, read in miRNA dataframes 
```{r}
apul_mi <- read.csv("../output/10-shortRNA-ShortStack-comparison/Apul_results_mature_named.csv", header = T) %>%
  dplyr::select(MajorRNA, given_miRNA_name) %>%
  dplyr::rename("Apul_MajorRNA" = "MajorRNA", "Apul_given_miRNA_name" = "given_miRNA_name")
peve_mi <- read.csv("../output/10-shortRNA-ShortStack-comparison/Peve_results_mature_named.csv", header = T) %>%
  dplyr::select(MajorRNA, given_miRNA_name) %>%
  dplyr::rename("Peve_MajorRNA" = "MajorRNA", "Peve_given_miRNA_name" = "given_miRNA_name")
pmea_mi <- read.csv("../output/10-shortRNA-ShortStack-comparison/Ptuh_results_mature_named.csv", header = T) %>%
  dplyr::select(MajorRNA, given_miRNA_name) %>%
  dplyr::rename("Pmea_MajorRNA" = "MajorRNA", "Pmea_given_miRNA_name" = "given_miRNA_name")
```

Join files
```{r}
join <- apul_mi %>%
  full_join(peve_mi, by = c("Apul_MajorRNA" = "Peve_MajorRNA"))
```

Doesn't work because the miR-2025 and miR-2036 are not exactly the same sequence...look back into this. 



### piRNAs

Javi wrote code for the piRNA [length comparison](https://github.com/urol-e5/deep-dive/blob/main/DEF-cross-species/code/12-piRNA-comparizon.Rmd). I'm going to modify his code.

Read in length data for piRNAs 
```{r}
APUL_L = list.files(path = "../../D-Apul/output/18-Apul-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

PMEA_L = list.files(path = "../../F-Pmea/output/18-Pmea-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

PEVE_L = list.files(path = "../../E-Peve/output/18-Peve-piRNA/0_piRNA_pipeline_proTRAC/piRNA_length_data", pattern=".txt", recursive = T, full.names = T)

L = c(APUL_L,PMEA_L,PEVE_L)
names <- L %>%
  substring(.,75) %>%
  gsub("_piRNA_length.txt", "", .)

DFs = lapply(L, function(x) {
  data <- read.table(paste0(x), comment.char = "")
  colnames(data) <- c("freq", "length")
  return(data)
})
```

Count piRNA lengths by species  
```{r}
combo = Reduce(function(...) merge(..., by = "length", all=T), DFs)
colnames(combo) <- c("length", names)
combo <- combo %>%
  mutate(ACR = rowSums(combo[,2:6])) %>%
  mutate(POC = rowSums(combo[,7:11])) %>%
  mutate(POR = rowSums(combo[,12:14])) %>%
  select(length,ACR,POC,POR) %>%
  pivot_longer(c("ACR", "POC", "POR"), values_to = "sums", names_to = "group") %>%
  mutate(Species = case_when(
    group == "ACR" ~ "A. pulchra", 
    group == "POR" ~ "P. evermanni",
    group == "POC" ~ "P. tuahiniensis"))
```

Plot piRNA lengths of all species
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_pi <- ggplot(combo, aes(x = length, y = sums, fill = Species)) + 
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  scale_y_continuous(expand = c(0, 0)) +   # Bars sit on x-axis
  scale_x_continuous(breaks = unique(combo$length)) +  # Ensure every "length" value appears on the x-axis
  labs(x = "Length (bp)", y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.title = element_text(size = 40, face = "bold"),  # Increase legend title size
    legend.text = element_text(size = 40),
    legend.position = "none"                              
  ) +
  guides(fill = guide_legend(override.aes = list(size = 20)));  # Increase the size of legend squares
length_plot_pi

ggsave(filename = "../figures/piRNA_lengths.png", length_plot_pi, width = 34, height = 20)
ggsave(filename = "../figures/piRNA_lengths.pdf", length_plot_pi, width = 34, height = 20)
```

### lncRNAs 

The actual overlap was calculated by Zach but I am just going to make a tile plot to match the miRNA one above. Similar to the miRNA tile plot above, I am not going to be adding colors or proportions. 
```{r}
# Create the data frame for the 3x3 grid
data <- data.frame(
  x = c("A. pulchra", "P. tuahiniensis", "P. evermanni", "A. pulchra", "P. tuahiniensis", 
        "P. evermanni", "A. pulchra", "P. tuahiniensis", "P. evermanni"),
  y = c("A. pulchra", "A. pulchra", "A. pulchra", "P. tuahiniensis", "P. tuahiniensis", 
        "P. tuahiniensis", "P. evermanni", "P. evermanni", "P. evermanni"),
  value = c(16206, 1799, 1763, 1799, 12395, 1044, 1763, 1044, 7378)
)

# Add a column to identify repetitive (lower triangle) comparisons
data <- data %>%
  mutate(
    repetitive = ifelse(x > y, "yes", "no"),    # Flag repetitive squares
    shared = ifelse(repetitive == "yes", NA, value)  # Assign NA to repetitive cells for color gradient
  )

# Convert x and y to factors to control order
#data$x <- factor(data$x, levels = c("A. pulchra", "P. tuahiniensis", "P. evermanni"))
#data$y <- factor(data$y, levels = c("A. pulchra", "P. tuahiniensis", "P. evermanni"))

tile_overlap_lnc <- ggplot(data, aes(x = x, y = y)) +
  geom_tile(aes(fill = repetitive), color = "black", size = 1) +               # Color tiles based on repetitive status
  geom_tile(data = subset(data, repetitive == "yes"),                # Overlay grey repetitive tiles
            fill = "gray", color = "black", size = 1) +
  geom_text(aes(label = ifelse(repetitive == "no", value, "")),      # Display values only for non-repetitive squares
            size = 25) +                                             # Add text in the middle of each tile
  #scale_fill_gradient(low = "blue", high = "red", na.value = "gray80") +  # Gradient for values, grey for NA
  scale_fill_manual(values = c("yes" = "gray80", "no" = "white"),    # Grey out repetitive squares, white for others
                    name = "Repetitive") +   
  theme_minimal() +
  theme(
    axis.title = element_blank(),                                    # Remove axis titles
    axis.text = element_text(size = 50, colour = "black", face = "italic"),              # Customize axis text
    axis.text.x = element_text(angle = 75, hjust = 1),  # Rotate and adjust label position
    #axis.line = element_line(size = 5, color = "black"),
    #panel.grid = element_blank(),                                    # Remove grid lines
    #panel.border = element_blank(),                                   # Remove panel borders
    legend.position = "none") +       
  coord_fixed(); tile_overlap_lnc

ggsave(filename = "../figures/lncRNA_tile_overlap.png", tile_overlap_lnc, width = 25, height = 35)
ggsave(filename = "../figures/lncRNA_tile_overlap.pdf", tile_overlap_lnc, width = 25, height = 35)
```

Not sure if this is correct...will check with Zach.

Calculate length from bed files
```{r}
apul_lnc_bed <- read.delim("../../D-Apul/output/05.33-lncRNA-discovery/Apul_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Acropora pulchra")

peve_lnc_bed <- read.delim("../../E-Peve/output/Peve_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Porites evermanni")

pmea_lnc_bed <- read.delim("../../F-Pmea/output/02-lncRNA-discovery/Pmea_lncRNA.bed", header = F) %>%
  rename("lncRNA" = "V1", "start" = "V2", "end" = "V3") %>%
  mutate(length = end-start) %>%
  mutate(Species = "Pocillopora tuahiniensis")

# Combine the data frames into one
all_lengths_lnc <- rbind(apul_lnc_bed, peve_lnc_bed, pmea_lnc_bed)
```

Define custom bins so that the lncRNAs are grouped into 200-300 bp, 300-400 bp, etc
```{r}
all_lengths_lnc <- all_lengths_lnc %>%
  mutate(length_bin = cut(
    length,
    breaks = c(200, 300, 400, 500, 600, 700, 800, 900, 1000, Inf),  # Define custom breaks
    labels = c("200-300", "300-400", "400-500", "500-600", "600-700", "700-800", "800-900", "900-1000", ">1000"),  # Custom labels for each bin
    include.lowest = TRUE,  # Include the lowest value in the first interval
    right = FALSE           # Make the intervals left-closed (e.g., 200-300 includes 200, but not 300)
  ))
```

Calculate frequency of lncRNAs in each bin by species
```{r}
length_freq <- all_lengths_lnc %>%
  group_by(Species, length_bin) %>%
  summarize(frequency = n(), .groups = 'drop')
```

Plot lncRNA lengths for all species 
```{r}
# Set our color scheme for plotting -- options for both the abbreviated labels or the full, correct species names
species_colors_nolabel <- c('#408EC6', '#1E2761', '#7A2048')

length_plot_lnc <- ggplot(length_freq, aes(x = length_bin, y = frequency, fill = Species)) +
  geom_bar(position = "dodge",
           stat = "identity",
           color = "black",  # Bar outline color
           size = 2,         # Thicker bar outlines
           width = 0.9) + 
  scale_fill_manual(values = species_colors_nolabel) + 
  # Use scale_x_discrete instead of scale_x_continuous
  scale_x_discrete(name = "Length (bp)") +  # Ensure every "length_bin" value appears on the x-axis
  scale_y_continuous(name = "Read count", expand = c(0, 0)) +  # Remove padding on y-axis so bars sit on x-axis
  labs(y = "Read count") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 60, face = "bold", margin = margin(t = 100, r = 100, b = 100, l = 100)),  # Space between titles and text
    axis.text = element_text(size = 55, colour = "black", margin = margin(t = 10, r = 10, b = 10, l = 10)),  # Axis text size
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis text for better readability
    axis.line = element_line(size = 1.5, color = "black"),  # Darker, thicker axis lines
    axis.ticks = element_line(size = 1.5, color = "black"), # Thicker axis ticks
    plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
    legend.title = element_text(size = 40, face = "bold"),  # Increase legend title size
    legend.text = element_text(size = 40),
    legend.position = "none"                              
  ) +
  guides(fill = guide_legend(override.aes = list(size = 20)));  # Increase the size of legend squares

length_plot_lnc

ggsave(filename = "../figures/lncRNA_lengths.png", length_plot_lnc, width = 34, height = 20)
ggsave(filename = "../figures/lncRNA_lengths.pdf", length_plot_lnc, width = 34, height = 20)
```


### ALL ncRNAs 

Still need lncRNA similarity data. Once I have that, I will put the lncRNA venn plot into the figure below 

Make dummy plot and extract legend so I can add it to main plot
```{r}
# Create a dummy data frame with the species and their associated colors
species_data <- data.frame(
  Species = c("A. pulchra", "P. tuahiniensis", "P. evermanni"),
  x = c("A", "B", "C"),  # Dummy x values
  y = 1:3   # Dummy y values
)

# Create a plot with the desired colors and species
plot <- ggplot(species_data, aes(x = x, y = y, fill = Species)) +
  geom_bar(position = "dodge", stat = "identity",            color = "black",  # Bar outline color
           size = 1,         # Thicker bar outlines
           width = 0.5) +
  scale_fill_manual(values = species_colors_nolabel) + 
  theme_void() +  # Remove plot background
  theme(legend.position = "right", 
         legend.key.size = unit(4, "cm"),  # Increase legend box size
         plot.margin = unit(c(1, 1, 1, 1), "cm"),                # Margin around plot (top, right, bottom, left)
         legend.title = element_text(size = 80, face = "bold"),  # Increase legend title size
         legend.spacing.y = unit(100, "cm"),  # Increase vertical spacing between legend boxes
         legend.text = element_text(size = 80, face = "italic"),); plot  

# Extract only the legend
legend <- cowplot::get_legend(plot)

# Draw the legend
grid.arrange(legend)
```

Plot RNA lengths + venns w/ legend 
```{r}
plot <- plot_grid(length_plot_lnc, tile_overlap_lnc, length_plot_mi, tile_overlap_mi, length_plot_pi, grid.arrange(legend), labels = c("A", "B", "C", "D", "E"), label_size = 80, nrow = 3, ncol = 2)

ggsave(filename = "../figures/ncRNA_lengths_overlap.png", plot, width = 55, height = 45, limitsize = F)
ggsave(filename = "../figures/ncRNA_lengths_overlap.pdf", plot, width = 55, height = 45, limitsize = F)
```




