---
title: "Closest gene comparisons across species"
author: "Jill Ashey"
date: "2024-09-11"
output: html_document
---

This code will examine the overlap (if any) between the GO enrichment of the closest gene across species. See the following rmds for TopGO analysis by species: XXXXXXXX

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Biostrings) # Biostrings_2.62.0 
require(ggplot2)    # ggplot2_3.3.6
library(ggpubr)
library(ggseqlogo)  # install.packages("ggseqlogo")
library(reshape2)   # reshape2_1.4.4
library(data.table)
library(gghighlight) #install.packages("gghighlight")
library(plotrix)
library(dplyr)
library(tidyr)
library(tibble)
library(tidyverse)
library(ggvenn)
```

## piRNA

Compare terms identified in piRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_pi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_piRNA.csv")
apul_go_pi$Species <- "Acropora pulchra"
peve_go_pi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_piRNA.csv")
peve_go_pi$Species <- "Porites evermanni"
pmea_go_pi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_piRNA.csv")
pmea_go_pi$Species <- "Pocillopora tuahiniensis"

all_go_pi <- rbind(apul_go_pi, pmea_go_pi, peve_go_pi)
```

Subset by ontology 
```{r}
# Apul
apul_go_pi_bp <- apul_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_pi_cc <- apul_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_pi_mf <- apul_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_pi_bp <- peve_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_pi_cc <- peve_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_pi_mf <- peve_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_pi_bp <- pmea_go_pi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_pi_cc <- pmea_go_pi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_pi_mf <- pmea_go_pi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_pi <- list(APUL=apul_go_pi_bp$GeneOntologyIDs, PTUH=pmea_go_pi_bp$GeneOntologyIDs,PEVE=peve_go_pi_bp$GeneOntologyIDs)

bp_venn_pi <- ggvenn(GOs_BP_wide, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_pi

# Cellular Components
GOs_CC_wide_pi <- list(APUL=apul_go_pi_cc$GeneOntologyIDs, PTUH=pmea_go_pi_cc$GeneOntologyIDs,PEVE=peve_go_pi_cc$GeneOntologyIDs)

cc_venn <- ggvenn(GOs_CC_wide, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_pi

# Molecular Functions
GOs_MF_wide_pi <- list(APUL=apul_go_pi_mf$GeneOntologyIDs, PTUH=pmea_go_pi_mf$GeneOntologyIDs,PEVE=peve_go_pi_mf$GeneOntologyIDs)

mf_venn_pi <- ggvenn(GOs_MF_wide, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_pi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_pi <- all_go_pi %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_pi_df <- all_go_pi %>%
  filter(GeneOntologyIDs %in% shared_go_pi$GeneOntologyIDs)

# Plot 
GO_plot_pi <- ggplot(shared_go_pi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_pi

ggsave(filename = "../figures/piRNA_shared_GOs_plot.png", GO_plot_pi, width = 25, height = 25)
ggsave(filename = "../figures/piRNA_shared_GOs_plot.pdf", GO_plot_pi, width = 25, height = 25)
```

I like this plot a lot!






## miRNA

Compare terms identified in miRNA GO enrichment analysis across species

Read in data from all species
```{r}
apul_go_mi <- read.csv("../../D-Apul/output/19-bedtools-closest/Apul_GO_en_sig_gene_miRNA.csv")
apul_go_mi$Species <- "Acropora pulchra"
peve_go_mi <- read.csv("../../E-Peve/output/19-bedtools-closest/Peve_GO_en_sig_gene_miRNA.csv")
peve_go_mi$Species <- "Porites evermanni"
pmea_go_mi <- read.csv("../../F-Pmea/output/19-bedtools-closest/Pmea_GO_en_sig_gene_miRNA.csv")
pmea_go_mi$Species <- "Pocillopora tuahiniensis"

all_go_mi <- rbind(apul_go_mi, pmea_go_mi, peve_go_mi)
```

Subset by ontology 
```{r}
# Apul
apul_go_mi_bp <- apul_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
apul_go_mi_cc <- apul_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
apul_go_mi_mf <- apul_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Peve
peve_go_mi_bp <- peve_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
peve_go_mi_cc <- peve_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
peve_go_mi_mf <- peve_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")

# Pmea
pmea_go_mi_bp <- pmea_go_mi %>%
  dplyr::filter(ontology == "Biological Processes")
pmea_go_mi_cc <- pmea_go_mi %>%
  dplyr::filter(ontology == "Cellular Components")
pmea_go_mi_mf <- pmea_go_mi %>%
  dplyr::filter(ontology == "Molecular Functions")
```

Look at overlaps with venn diagrams 
```{r}
# Biological Processes
GOs_BP_wide_mi <- list(APUL=apul_go_mi_bp$GeneOntologyIDs, PTUH=pmea_go_mi_bp$GeneOntologyIDs,PEVE=peve_go_mi_bp$GeneOntologyIDs)

bp_venn_mi <- ggvenn(GOs_BP_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); bp_venn_mi

# Cellular Components
GOs_CC_wide_mi <- list(APUL=apul_go_mi_cc$GeneOntologyIDs, PTUH=pmea_go_mi_cc$GeneOntologyIDs,PEVE=peve_go_mi_cc$GeneOntologyIDs)

cc_venn_mi <- ggvenn(GOs_CC_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); cc_venn_mi

# Molecular Functions
GOs_MF_wide_mi <- list(APUL=apul_go_mi_mf$GeneOntologyIDs, PTUH=pmea_go_mi_mf$GeneOntologyIDs,PEVE=peve_go_mi_mf$GeneOntologyIDs)

mf_venn_mi <- ggvenn(GOs_MF_wide_mi, fill_color = c("#408EC6", "#7A2048", "#1E2761"),
  stroke_size = 0.5, set_name_size = 0, show_percentage = F, stroke_color = F); mf_venn_mi
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go <- all_go %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_df <- all_go %>%
  filter(GeneOntologyIDs %in% shared_go$GeneOntologyIDs)

# Plot 
GO_plot <- ggplot(shared_go_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 4, color = "black") +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('Enrichment score') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  coord_flip(); GO_plot
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_go_mi <- all_go_mi %>%
  group_by(GeneOntologyIDs) %>%  # Group by both GeneOntologyIDs and ontology
  summarise(unique_species = n_distinct(Species), .groups = 'keep') %>%  # Count distinct species for each GO term and ontology
  filter(unique_species > 1)

shared_go_mi <- all_go_mi %>%
  group_by(GeneOntologyIDs) %>%
  filter(if (any(ontology == "Cellular Components")) ontology == "Cellular Components" else TRUE) %>%
  summarise(unique_species = n_distinct(Species), .groups = 'drop') %>%
  filter(unique_species > 1)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_go_mi_df <- all_go_mi %>%
  filter(GeneOntologyIDs %in% shared_go_mi$GeneOntologyIDs)

# Plot 
GO_plot_mi <- ggplot(shared_go_mi_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 4, color = "black") +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('Enrichment score') +
  facet_grid(vars(ontology)) + 
  theme_bw() +
  coord_flip(); GO_plot_mi

ggsave(filename = "../figures/miRNA_shared_GOs_plot.png", GO_plot_mi, width = 15, height = 20)
```

Weird...having issues with this plot. It is plotting terms that are unique only to one species when I am asking it to plot only terms that are shared by 2+ species. When I look in the df however, the terms actually are shared, the plot itself just isn't plotting them. Ah it might be because some GO terms belong to multiple ontologies...which is dumb. It looks like most of the issue is arising from POC. I have tried to remove terms that are associated with 1+ ontology and only keep the row that is associated with cellular components but this has not succeeded. UGH


