---
title: "Species comparison of miRNA target enrichment"
author: "Jill Ashey"
date: "2024-10-30"
output: html_document
---

This code will examine the overlap (if any) between the GO enrichment of the miRNA-targeted genes across species.

## All miRNAs 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Biostrings) # Biostrings_2.62.0 
require(ggplot2)    # ggplot2_3.3.6
library(ggpubr)
library(ggseqlogo)  # install.packages("ggseqlogo")
library(reshape2)   # reshape2_1.4.4
library(data.table)
library(gghighlight) #install.packages("gghighlight")
library(plotrix)
library(dplyr)
library(tidyr)
library(tibble)
library(tidyverse)
library(ggvenn)
library(cowplot)
library(gridExtra)
library(ComplexUpset)
```

Read in data 
```{r}
apul_target <- read.csv("../../D-Apul/output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_miRNA_targets.csv")
apul_target$Species <- "Acropora pulchra"
peve_target <- read.csv("../../E-Peve/output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_miRNA_targets.csv")
peve_target$Species <- "Porites evermanni"
pmea_target <- read.csv("../../F-Pmea/output/21-Pmea-miRNA-target-prediction/Pmea_GO_en_sig_miRNA_targets.csv")
pmea_target$Species <- "Pocillopora tuahiniensis"

all_target <- rbind(apul_target, peve_target, pmea_target)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_target <- all_target %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_target_df <- all_target %>%
  filter(GeneOntologyIDs %in% shared_target$GeneOntologyIDs)

# Save as csv 
write.csv(shared_target_df, "../output/21-miRNA-target-prediction/miRNA_shared_target_go_by_species.csv")

# Plot 
GO_plot_target <- ggplot(shared_target_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_target

ggsave(filename = "../figures/miRNA_targets_GOs_plot.png", GO_plot_target, width = 25, height = 25)
ggsave(filename = "../figures/miRNA_targets_GOs_plot.pdf", GO_plot_target, width = 25, height = 25)
```

Plot only BP 
```{r}
shared_target_df_bp <- shared_target_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_BP_plot <- ggplot(shared_target_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  #facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_BP_plot

ggsave(filename = "../figures/miRNA_targets_GO_BP_plot.png", GO_BP_plot, width = 40, height = 25)
ggsave(filename = "../figures/miRNA_targets_GO_BP_plot.pdf", GO_BP_plot, width = 40, height = 25)
```

## mir-100

Read in data 
```{r}
apul_mir100_target <- read.csv("../../D-Apul/output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_mir-100_targets.csv")
apul_mir100_target$Species <- "Acropora pulchra"
peve_mir100_target <- read.csv("../../E-Peve/output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_mir-100_targets.csv")
peve_mir100_target$Species <- "Porites evermanni"
pmea_mir100_target <- read.csv("../../F-Pmea/output/21-Pmea-miRNA-target-prediction/Pmea_GO_en_sig_mir-100_targets.csv")
pmea_mir100_target$Species <- "Pocillopora tuahiniensis"

all_mir100_target <- rbind(apul_mir100_target, peve_mir100_target, pmea_mir100_target)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_mir100_target <- all_mir100_target %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_mir100_target_df <- all_mir100_target %>%
  filter(GeneOntologyIDs %in% shared_mir100_target$GeneOntologyIDs)

# Plot 
GO_plot_mir100_target <- ggplot(shared_mir100_target_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_mir100_target

ggsave(filename = "../figures/mir-100_targets_GO_plot.png", GO_plot_mir100_target, width = 40, height = 25)
ggsave(filename = "../figures/mir-100_targets_GO_plot.pdf", GO_plot_mir100_target, width = 40, height = 25)
```

Plot only BP 
```{r}
shared_mir100_target_df_bp <- shared_mir100_target_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_mir100_BP_plot <- ggplot(shared_mir100_target_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_BP_plot

ggsave(filename = "../figures/mir-100_targets_GO_BP_plot.png", GO_mir100_BP_plot, width = 40, height = 25)
ggsave(filename = "../figures/mir-100_targets_GO_BP_plot.pdf", GO_mir100_BP_plot, width = 40, height = 25)
```

## mir-2023

Read in data 
```{r}
apul_mir2023_target <- read.csv("../../D-Apul/output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_mir-2023_targets.csv")
apul_mir2023_target$Species <- "Acropora pulchra"
peve_mir2023_target <- read.csv("../../E-Peve/output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_mir-2023_targets.csv")
peve_mir2023_target$Species <- "Porites evermanni"
pmea_mir2023_target <- read.csv("../../F-Pmea/output/21-Pmea-miRNA-target-prediction/Pmea_GO_en_sig_mir-2023_targets.csv")
pmea_mir2023_target$Species <- "Pocillopora tuahiniensis"

all_mir2023_target <- rbind(apul_mir2023_target, peve_mir2023_target, pmea_mir2023_target)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_mir2023_target <- all_mir2023_target %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_mir2023_target_df <- all_mir2023_target %>%
  filter(GeneOntologyIDs %in% shared_mir2023_target$GeneOntologyIDs)

# Plot 
GO_plot_mir2023_target <- ggplot(shared_mir2023_target_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_mir2023_target

ggsave(filename = "../figures/mir-2023_targets_GO_plot.png", GO_plot_mir2023_target, width = 40, height = 25)
ggsave(filename = "../figures/mir-2023_targets_GO_plot.pdf", GO_plot_mir2023_target, width = 40, height = 25)
```

Plot only BP 
```{r}
shared_mir2023_target_df_bp <- shared_mir2023_target_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_mir2023_BP_plot <- ggplot(shared_mir2023_target_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_mir2023_BP_plot

ggsave(filename = "../figures/mir-2023_targets_GO_BP_plot.png", GO_mir2023_BP_plot, width = 40, height = 25)
ggsave(filename = "../figures/mir-2023_targets_GO_BP_plot.pdf", GO_mir2023_BP_plot, width = 40, height = 25)
```

## mir-2025

Read in data 
```{r}
apul_mir2025_target <- read.csv("../../D-Apul/output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_mir-2025_targets.csv")
apul_mir2025_target$Species <- "Acropora pulchra"
peve_mir2025_target <- read.csv("../../E-Peve/output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_mir-2025_targets.csv")
peve_mir2025_target$Species <- "Porites evermanni"
pmea_mir2025_target <- read.csv("../../F-Pmea/output/21-Pmea-miRNA-target-prediction/Pmea_GO_en_sig_mir-2025_targets.csv")
pmea_mir2025_target$Species <- "Pocillopora tuahiniensis"

all_mir2025_target <- rbind(apul_mir2025_target, peve_mir2025_target, pmea_mir2025_target)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_mir2025_target <- all_mir2025_target %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_mir2025_target_df <- all_mir2025_target %>%
  filter(GeneOntologyIDs %in% shared_mir2025_target$GeneOntologyIDs)

# Plot 
GO_plot_mir2025_target <- ggplot(shared_mir2025_target_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_mir2025_target

ggsave(filename = "../figures/mir-2025_targets_GO_plot.png", GO_plot_mir2025_target, width = 40, height = 25)
ggsave(filename = "../figures/mir-2025_targets_GO_plot.pdf", GO_plot_mir2025_target, width = 40, height = 25)
```

Plot only BP 
```{r}
shared_mir2025_target_df_bp <- shared_mir2025_target_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_mir2025_BP_plot <- ggplot(shared_mir2025_target_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_mir2025_BP_plot

ggsave(filename = "../figures/mir-2025_targets_GO_BP_plot.png", GO_mir2025_BP_plot, width = 40, height = 25)
ggsave(filename = "../figures/mir-2025_targets_GO_BP_plot.pdf", GO_mir2025_BP_plot, width = 40, height = 25)
```

## mir-2036

Read in data 
```{r}
apul_mir2036_target <- read.csv("../../D-Apul/output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_mir-2036_targets.csv")
apul_mir2036_target$Species <- "Acropora pulchra"
peve_mir2036_target <- read.csv("../../E-Peve/output/21-Peve-miRNA-target-prediction/Peve_GO_en_sig_mir-2036_targets.csv")
peve_mir2036_target$Species <- "Porites evermanni"
pmea_mir2036_target <- read.csv("../../F-Pmea/output/21-Pmea-miRNA-target-prediction/Pmea_GO_en_sig_mir-2036_targets.csv")
pmea_mir2036_target$Species <- "Pocillopora tuahiniensis"

all_mir2036_target <- rbind(apul_mir2036_target, peve_mir2036_target, pmea_mir2036_target)
```

Assess shared GO terms between species 
```{r}
# Count the number of unique species for each GeneOntologyID
shared_mir2036_target <- all_mir2036_target %>%
  group_by(GeneOntologyIDs) %>%
  summarise(unique_species = n_distinct(Species)) %>%
  filter(unique_species > 1) %>%
  dplyr::select(GeneOntologyIDs)

# Create a new dataframe with only the shared GeneOntologyIDs
shared_mir2036_target_df <- all_mir2036_target %>%
  filter(GeneOntologyIDs %in% shared_mir2036_target$GeneOntologyIDs)

# Plot 
GO_plot_mir2036_target <- ggplot(shared_mir2036_target_df, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  theme_bw() +
  theme(
    axis.title = element_text(size = 36, face = "bold"),        # Axis title size
    axis.text = element_text(size = 34, colour = "black"),                        # Axis text size
    legend.title = element_text(size = 34, face = "bold"),      # Legend title size
    legend.text = element_text(size = 32),                      # Legend text size
    strip.text = element_text(size = 34, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside"    
  ) +
  coord_flip(); GO_plot_mir2036_target

ggsave(filename = "../figures/mir-2036_targets_GO_plot.png", GO_plot_mir2036_target, width = 40, height = 25)
ggsave(filename = "../figures/mir-2036_targets_GO_plot.pdf", GO_plot_mir2036_target, width = 40, height = 25)
```

Plot only BP 
```{r}
shared_mir2036_target_df_bp <- shared_mir2036_target_df %>%
  dplyr::filter(ontology == "Biological Processes")
  
# Plot 
GO_mir2036_BP_plot <- ggplot(shared_mir2036_target_df_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(ontology), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_mir2036_BP_plot

ggsave(filename = "../figures/mir-2036_targets_GO_BP_plot.png", GO_mir2036_BP_plot, width = 40, height = 25)
ggsave(filename = "../figures/mir-2036_targets_GO_BP_plot.pdf", GO_mir2036_BP_plot, width = 40, height = 25)
```

## Shared miRNAs 

Join shared miRNAs into one df for BP
```{r}
shared_mirna_target_bp <- rbind(shared_mir100_target_df_bp, shared_mir2023_target_df_bp, shared_mir2025_target_df_bp, shared_mir2036_target_df_bp)

# Remove species info before miRNA name
shared_mirna_target_bp$mirna <- gsub("apul-", "", shared_mirna_target_bp$mirna)
shared_mirna_target_bp$mirna <- gsub("peve-", "", shared_mirna_target_bp$mirna)
shared_mirna_target_bp$mirna <- gsub("ptuh-", "", shared_mirna_target_bp$mirna)

# Save as csv 
write.csv(shared_mirna_target_bp, "../output/21-miRNA-target-prediction/conserved_miRNA_shared_target_go_by_species_BP.csv")
```

Plot 
```{r}
shared_GO_BP_plot <- ggplot(shared_mirna_target_bp, aes(x = Term, y = -log10(Fisher), shape = Species, fill = Species)) +
  geom_point(size = 25, color = "black", stroke = 2) +  # Black border around shapes
  scale_shape_manual(values = c(21, 22, 23)) +  # Shapes that can be filled
  scale_fill_manual(values = c("#408EC6", "#7A2048", "#1E2761")) +  # Custom colors for each species
  #scale_size(range = c(2, 12)) + 
  xlab('') + 
  ylab('log(Enrichment score)') +
  facet_grid(vars(mirna), scales = "free", space = "free_y") + 
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 51, face = "bold"),        # Axis title size
    axis.text = element_text(size = 49, colour = "black"),                        # Axis text size
    axis.text.x = element_text(hjust = 1),  # Rotate and adjust label position
    legend.title = element_text(size = 44, face = "bold"),      # Legend title size
    legend.text = element_text(size = 42, face = "italic"),                      # Legend text size
    strip.text = element_text(size = 49, face = "bold"),        # Facet text size
    strip.background = element_rect(fill = "lightgray", color = "black", size = 1.5),         # Facet background
    axis.line = element_line(size = 1, colour = "black"),       # Enhanced axis lines
    axis.ticks = element_line(size = 1),                        # Thicker axis ticks
     #panel.border = element_blank()                             # Remove panel border
    #panel.grid.major = element_blank(),                         # Remove major grid lines
    #panel.grid.minor = element_blank()                          # Remove minor grid lines
        panel.border = element_rect(color = "black", size = 1.2),   # Enhanced facet border lines
    panel.grid.major = element_line(size = 0.5, color = "gray"), # Grid lines inside the facets
    panel.spacing = unit(1, "lines"),                           # Increase space between facets
    strip.placement = "outside",
    plot.margin = margin(t = 10, r = 10, b = 20, l = 10)        # Add space below the plot
  ); GO_BP_plot

ggsave(filename = "../figures/shared_miRNA_targets_GO_BP_plot.png", shared_GO_BP_plot, width = 40, height = 40)
ggsave(filename = "../figures/shared_miRNA_targets_GO_BP_plot.pdf", shared_GO_BP_plot, width = 40, height = 40)
```

