---
title: "pverr GFF fixing"
author: "jillashey"
date: "2023-04-03"
output: html_document
---

This script add transcript and gene id into GFF file for alignment.  

Here, I'll be adding transcript_id= and gene_id= to 'gene' column in order to properly assemble our aligned data  

Load libraries and data. 
```{r}
#Load libraries
library(tidyverse)
library(R.utils)
```

Load gene gff file 
```{r}
gff <- read.csv(file = "~/Desktop/GFFs/pverr/Pver_genome_assembly_v1.0.gff3", header = F, sep = "\t", skip = 1)
```

Rename columns 
```{r}
colnames(gff) <- c("scaffold", "Gene.Predict", "id", "gene.start","gene.stop", "pos1", "pos2","pos3", "gene")

# Remove all rows with "#" character - this gff has # denoting protein sequences. Since we don't need the protein sequences right now, I'm going to remove them 
gff <- gff[!grepl("#", gff$scaffold),]
```

Create transcript ID
```{r}
gff$transcript_id <- sub(";.*", "", gff$gene)
gff$transcript_id <- gsub("ID=", "", gff$transcript_id) #remove ID= 
head(gff)
```

Create Parent ID
```{r}
gff$parent_id <- sub(".*Parent=", "", gff$gene)
gff$parent_id <- sub(";.*", "", gff$parent_id)
gff$parent_id <- gsub("ID=", "", gff$parent_id) #remove ID= 
head(gff)
```

Add these values back into the gene column separated by semicolons
```{r}
gff <- gff %>% 
  mutate(gene = ifelse(id != "gene", paste0(gene, ";transcript_id=", gff$transcript_id, ";gene_id=", gff$parent_id),  paste0(gene)))
head(gff)

## if this gff does not work in the stringtie script, go back and edit so that transcript_id=Pver_g1.t2 instead of transcript_id=Pver_g1.t2.utr5p1 or transcript_id=Pver_g1.t2.exon1, etc
```

Save file 
```{r}
write.table(gff, file = "~/Desktop/GFFs/pverr/Pver_genome_assembly_v1_fixed.gff3", sep="\t", col.names = FALSE, row.names=FALSE, quote=FALSE)
```

Zip file and upload to HPC for bioinformatic use

`gzip /Users/jillashey/Desktop/GFFs/pverr/Pver_genome_assembly_v1_fixed.gff3`

`scp /Users/jillashey/Desktop/GFFs/pverr/Pver_genome_assembly_v1_fixed.gff3.gz jillashey@ssh3.hac.uri.edu:/data/putnamlab/ashuffmyer/e5-deepdive/rna-seq/refs`

