---
title: "05.3-Apul-lncRNA-discovery"
author: "Steven Roberts"
date: "`r format(Sys.time(), '%d %B, %Y')`"  
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(DT)
library(Biostrings)
library(tm)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```



# Run HiSat on RNA-seq

## Grab Trimmed RNA-seq Reads

```{r, engine='bash'}

wget -r \
--no-directories --no-parent \
-P ../data/fastq/ \
-A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/

```
## Genome

```{r, engine='bash'}
cd ../data

curl -O http://gannet.fish.washington.edu/seashell/snaps/GCF_013753865.1_Amil_v2.1_genomic.fna
```


## HiSat

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2_extract_exons.py \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
> ../output/05.32-lncRNA-discovery/m_exon.tab
```

```{r, engine='bash', eval=TRUE}
head ../output/05.32-lncRNA-discovery/m_exon.tab
```

```{r, engine='bash'}

/home/shared/hisat2-2.2.1/hisat2_extract_splice_sites.py \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
> ../output/05.32-lncRNA-discovery/m_splice_sites.tab

```


```{r, engine='bash', eval=TRUE}
head ../output/05.32-lncRNA-discovery/m_splice_sites.tab
```

```{r, engine='bash'}
/home/shared/hisat2-2.2.1/hisat2-build \
../data/GCF_013753865.1_Amil_v2.1_genomic.fna \
../data/GCF_013753865.1_Amil_v2.1.index \
--exon ../output/05.32-lncRNA-discovery/m_exon.tab \
--ss ../output/05.32-lncRNA-discovery/m_splice_sites.tab \
-p 40 \
../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf \
2> ../output/05.32-lncRNA-discovery/hisat2-build_stats.txt

```

```{r, engine='bash', eval=TRUE}
find ../data/fastq/*gz
```

```{r, engine='bash', eval=TRUE}
find ../data/fastq/*R2_001.fastp-trim.20230519.fastq.gz \
| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
echo {}
```

```{r, engine='bash'}
find ../data/fastq/*R2_001.fastp-trim.20230519.fastq.gz \
| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
/home/shared/hisat2-2.2.1/hisat2 \
-x ../data/GCF_013753865.1_Amil_v2.1.index \
-p 20 \
-1 ../data/fastq/{}-S1-TP2_R1_001.fastp-trim.20230519.fastq.gz \
-2 ../data/fastq/{}-S1-TP2_R2_001.fastp-trim.20230519.fastq.gz \
-S ../output/05.32-lncRNA-discovery/{}.sam
```

```
47710408 reads; of these:
  47710408 (100.00%) were paired; of these:
    27056517 (56.71%) aligned concordantly 0 times
    19229695 (40.31%) aligned concordantly exactly 1 time
    1424196 (2.99%) aligned concordantly >1 times
    ----
    27056517 pairs aligned concordantly 0 times; of these:
      1316579 (4.87%) aligned discordantly 1 time
    ----
    25739938 pairs aligned 0 times concordantly or discordantly; of these:
      51479876 mates make up the pairs; of these:
        41152624 (79.94%) aligned 0 times
        9290398 (18.05%) aligned exactly 1 time
        1036854 (2.01%) aligned >1 times
56.87% overall alignment rate
42864294 reads; of these:
  42864294 (100.00%) were paired; of these:
    23649791 (55.17%) aligned concordantly 0 times
    17703533 (41.30%) aligned concordantly exactly 1 time
    1510970 (3.53%) aligned concordantly >1 times
    ----
    23649791 pairs aligned concordantly 0 times; of these:
      1250605 (5.29%) aligned discordantly 1 time
    ----
    22399186 pairs aligned 0 times concordantly or discordantly; of these:
      44798372 mates make up the pairs; of these:
        35369621 (78.95%) aligned 0 times
        8218798 (18.35%) aligned exactly 1 time
        1209953 (2.70%) aligned >1 times
58.74% overall alignment rate
43712298 reads; of these:
  43712298 (100.00%) were paired; of these:
    30355359 (69.44%) aligned concordantly 0 times
    12131465 (27.75%) aligned concordantly exactly 1 time
    1225474 (2.80%) aligned concordantly >1 times
    ----
    30355359 pairs aligned concordantly 0 times; of these:
      848325 (2.79%) aligned discordantly 1 time
    ----
    29507034 pairs aligned 0 times concordantly or discordantly; of these:
      59014068 mates make up the pairs; of these:
        51031599 (86.47%) aligned 0 times
        6695839 (11.35%) aligned exactly 1 time
        1286630 (2.18%) aligned >1 times
41.63% overall alignment rate
47501524 reads; of these:
  47501524 (100.00%) were paired; of these:
    27827915 (58.58%) aligned concordantly 0 times
    18298532 (38.52%) aligned concordantly exactly 1 time
    1375077 (2.89%) aligned concordantly >1 times
    ----
    27827915 pairs aligned concordantly 0 times; of these:
      1321896 (4.75%) aligned discordantly 1 time
    ----
    26506019 pairs aligned 0 times concordantly or discordantly; of these:
      53012038 mates make up the pairs; of these:
        42654805 (80.46%) aligned 0 times
        9174277 (17.31%) aligned exactly 1 time
        1182956 (2.23%) aligned >1 times
55.10% overall alignment rate
42677752 reads; of these:
  42677752 (100.00%) were paired; of these:
    25646587 (60.09%) aligned concordantly 0 times
    15654238 (36.68%) aligned concordantly exactly 1 time
    1376927 (3.23%) aligned concordantly >1 times
    ----
    25646587 pairs aligned concordantly 0 times; of these:
      1105089 (4.31%) aligned discordantly 1 time
    ----
    24541498 pairs aligned 0 times concordantly or discordantly; of these:
      49082996 mates make up the pairs; of these:
        38229072 (77.89%) aligned 0 times
        9393012 (19.14%) aligned exactly 1 time
        1460912 (2.98%) aligned >1 times
55.21% overall alignment rate
```


```{r, engine='bash'}
for samfile in ../output/05.32-lncRNA-discovery/*.sam; do
  bamfile="${samfile%.sam}.bam"
  sorted_bamfile="${samfile%.sam}.sorted.bam"
  
  # Convert SAM to BAM
  /home/shared/samtools-1.12/samtools view -bS "$samfile" > "$bamfile"
  
  # Sort BAM
  /home/shared/samtools-1.12/samtools sort "$bamfile" -o "$sorted_bamfile"
  
  # Index sorted BAM
  /home/shared/samtools-1.12/samtools index "$sorted_bamfile"
done
```

# Samtools
Convert SAM files to Binary Alignment Map (BAM) files, then sorts and saves them in the $APUL_DIR/push/bam directory. BAM files save on storage and speed up the next step in processing.

# StringTie
StringTie uses the sorted BAM files to assemble transcripts for each sample, outputting them as GTF (Gene Transfer Format) files. And then merges all individual GTF assemblies into a single merged GTF file. This step extracts transcript information and merges GTFs from all samples--an important step in creating a canonical list of lncRNAs across all samples included in the pipeline.


Merges all individual GTF assemblies into a single merged GTF file.

```{r, engine='bash'}
find ../output/05.32-lncRNA-discovery/*sorted.bam \
| xargs basename -s .sorted.bam | xargs -I{} \
/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
-p 8 \
-G ../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff \
-o ../output/05.32-lncRNA-discovery/{}.gtf \
../output/05.32-lncRNA-discovery/{}.sorted.bam
```

```{r, engine='bash', eval=TRUE}
wc -l ../output/05.32-lncRNA-discovery/RNA*.gtf
ls ../output/05.32-lncRNA-discovery/RNA*.gtf
```

```{r, engine='bash'}
/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
--merge \
-G ../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff \
-o ../output/05.32-lncRNA-discovery/stringtie_merged.gtf \
../output/05.32-lncRNA-discovery/*.gtf
```

```{r, engine='bash', eval=TRUE}

wc -l ../output/05.32-lncRNA-discovery/stringtie_merged.gtf
head ../output/05.32-lncRNA-discovery/stringtie_merged.gtf


```




```{r, engine='bash'}
/home/shared/gffcompare-0.12.6.Linux_x86_64/gffcompare \
-r ../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff \
-G \
-o ../output/05.32-lncRNA-discovery/gffcompare_merged \
../output/05.32-lncRNA-discovery/stringtie_merged.gtf
```


```{r, engine='bash', eval=TRUE}
ls ../output/05.32-lncRNA-discovery/gffcompare_merged*
```

```{r, engine='bash', eval=TRUE}
head -3 ../output/05.32-lncRNA-discovery/gffcompare_merged.annotated.gtf
```




# Filter
Filters the combined GTF output from GFFcompare to select only the lines representing "transcripts" and excluding lines starting with "#" (these are lines in the output format from GFFcompare that don't contain transcript information). This step further filters for a class code of "u", and keep only those with lengths greater than 199 bases. The "u' class code from the GFFcompare step is for "unknown" transcripts, that is those that are not previously annotated in our reference GFF as protein coding. The size filter of +200nt is a common filtering step for isolating lncRNAs.

```{r, engine='bash'}
awk '$3 == "transcript" && $1 !~ /^#/ {print}' \
../output/05.32-lncRNA-discovery/gffcompare_merged.annotated.gtf | grep 'class_code "u"' | awk '$5 - $4 > 199 {print}' > ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.gtf
```

```{r, engine='bash', eval=TRUE}
head ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.gtf
```

# Bedtools
Extracts the sequence data from the `$FASTA` reference file based on the coordinates from the filtered GTF. The resulting sequences represent potential lncRNA candidates.

```{r, engine='bash'}
/home/shared/bedtools2/bin/fastaFromBed \
-fi ../data/GCF_013753865.1_Amil_v2.1_genomic.fna \
-bed ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.gtf \
-fo ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.fasta \
-name -split
```

```{r, engine='bash'}
fgrep -c ">" ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.fasta
head ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.fasta
```



# CPC2
Initializes a conda environment (Anaconda) and runs CPC2, a software to predict whether a transcript is coding or non-coding. The results are saved to the $OUTPUT_DIR. CPC2 uses ORF (Open Reading Frame) Analysis, Isometric Feature Mapping (Isomap), Sequence Homology, RNA Sequence Features, and Quality of Translation to assess coding potential and flag any transcripts we would want to exclude using the FASTA generated in the previous step.

```{r, engine='bash'}
eval "$(/opt/anaconda/anaconda3/bin/conda shell.bash hook)"
python /home/shared/CPC2_standalone-1.0.1/bin/CPC2.py \
-i ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.fasta \
-o ../output/05.32-lncRNA-discovery/Apul_CPC2
```


```{r, engine='bash', eval=TRUE}
wc -l head ../output/05.32-lncRNA-discovery/Apul_CPC2.txt
head ../output/05.32-lncRNA-discovery/Apul_CPC2.txt
```

#Filter
Filters the CPC2 results to get only noncoding transcripts (using the class "noncoding" from the CPC2 results) and extracts their IDs and matches these IDs with the sequences from the previous step to generate a GTF of long noncoding transcripts.

Matches these IDs with the sequences from the previous step to generate a GTF of noncoding transcripts.

```{r, engine='bash'}
awk '$8 == "noncoding" {print $1}' ../output/05.32-lncRNA-discovery/Apul_CPC2.txt > ../output/05.32-lncRNA-discovery/Apul_noncoding_transcripts_ids.txt

grep -Fwf ../output/05.32-lncRNA-discovery/Apul_noncoding_transcripts_ids.txt ../output/05.32-lncRNA-discovery/Apul_lncRNA_candidates.fasta > ../output/05.32-lncRNA-discovery/Apul_lncRNAs.gtf
```


```{r, engine='bash', eval=TRUE}
wc -l ../output/05.32-lncRNA-discovery/Apul_lncRNAs.gtf
head ../output/05.32-lncRNA-discovery/Apul_lncRNAs.gtf
```

# De-duplicate GTF
Removes duplicate entries from the GTF file. This results in a list of lncRNAs with transcript IDs that show scaffold and base range.
awk '!seen[$0]++' $OUTPUT_DIR/apul_merged_final_lncRNAs.gtf > $OUTPUT_DIR/apul_deduplicated_final_lncRNAs.gtf

# Reformat to BED
Converts the GTF format to BED (Browser Extensible Data) format, a commonly used format for representing genomic intervals. This allows us to use bedtools to generate a FASTA of lncRNAs.
awk -F":|-" '{print $3 "\t" $4 "\t" $5}' $OUTPUT_DIR/apul_deduplicated_final_lncRNAs.gtf > $OUTPUT_DIR/apul_deduplicated_final_lncRNAs.bed

# BEDTools
This extracts sequences from the reference FASTA based on the BED file's coordinates. The resulting sequences represent the final set of non-redundant lncRNAs.
$BEDTOOLS_PATH/bedtools \
getfasta -fi $FASTA -bed $OUTPUT_DIR/apul_deduplicated_final_lncRNAs.bed -fo $OUTPUT_DIR/apul_bedtools_lncRNAs.fasta -name
```

The results of this pipleine provide the user with GTF files containing transcript IDs (scaffold and sequence range), BED files displaying the same transcript IDs in BED format for use with BEDtools, and FASTA files containing the sequence of identified lncRNAs.