---
title: "TE_in_piRNA_clusters"
author: "Javier Rodriguez Casariego"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
knitr::opts_knit$set(root.dir = '/Users/jarcasariego/Dropbox/Other_projects/E5_work/deep-dive/D-Apul/')
```


```{r}
#This script generates stacked barplots comparing TE content in the entire genome to that of piRNA clusters
#load libraries
library(dplyr)   
library(ggplot2)
library(tidyverse)

my.outprefix <- paste(Sys.Date(),"Cluster",sep="_")

#read in the outfile from repeatmasker
out_file <- read.table("data/Amil/GCF_013753865.1_Amil_v2.1_genomic.fna.out", comment.char = "", header = F, fill = T, skip = 2)

#retains only 1)chromosome 2)start 3)stop 4)TE name 5)Family Name and 6)percent divergence
bed_file <- out_file[,c(5:7,11,2)]

#Assess how the families breakdown for the whole genome
#remove empty lines 
bed_file <- bed_file %>%
  drop_na(.)

#output bed file
write.table(bed_file, file = "output/18-Apul-piRNA/2_TE_piRNA_cluster_overlap/APUL_repeatmasker_bedfile.bed", sep = "\t", col.names = F, row.names = F, quote = F)

#retain only major family name and combine unclear and Unknown
bed_file$V11 <- factor(bed_file$V11)
bed_file$V11 <- gsub("/.*","",bed_file$V11)
bed_file$V11 <- gsub("unclear","Unknown",bed_file$V11)

#make a table of the families of TEs. How frequently do the different TE families appear in the genome?
table_of_TEs <- as.data.frame(table(bed_file$V11))

#remove simple repeats and unify "unknown" and "unclear"
table_of_TEs <- table_of_TEs[(!grepl("Low_complexity", table_of_TEs$Var1)),]
table_of_TEs <- table_of_TEs[(!grepl("Simple_repeat", table_of_TEs$Var1)),]

#create matrix with TE family frequency in the genome
cols.g <- rainbow(nrow(table_of_TEs))
table_of_TEs$percent = round(100*table_of_TEs$Freq/sum(table_of_TEs$Freq), digits = 1)
table_of_TEs$label = paste(table_of_TEs$Var1," (", table_of_TEs$percent,"%)", sep = "")
```

#Assess how the families breakdown for piRNA clusters
```{r, engine='bash'}
# Intersect cluster bedfile with TE bedfile
bedtools intersect -wo -a APUL_merged_cluster.bed -b APUL_repeatmasker_bedfile.bed > APUL_piRNA_cluster_TEs_intersect.txt

```

```{r}
#read in the output of bedtools intersect
cluster_TEs <- read.table("output/18-Apul-piRNA/0_piRNA_pipeline_proTRAC/APUL_piRNA_clusters_TEs_intersect.txt")

#repeat analysis as above with piRNA cluster TEs
cluster_TEs$V8 <- gsub("/.*","",cluster_TEs$V8)
cluster_TEs$V8 <- gsub("unclear","Unknown",cluster_TEs$V8)
table_of_cluster_TEs <- as.data.frame(table(cluster_TEs$V8))
table_of_cluster_TEs <- table_of_cluster_TEs[(!grepl("Low_complexity", table_of_cluster_TEs$Var1)),]
table_of_cluster_TEs <- table_of_cluster_TEs[(!grepl("Simple_repeat", table_of_cluster_TEs$Var1)),]
cols <- rainbow(nrow(table_of_cluster_TEs))
table_of_cluster_TEs$percent = round(100*table_of_cluster_TEs$Freq/sum(table_of_cluster_TEs$Freq), digits = 1)
table_of_cluster_TEs$label = paste(table_of_cluster_TEs$Var1," (", table_of_cluster_TEs$percent,"%)", sep = "")

#generate stacked barplot showing TE distribution of clusters vs whole genome
table_of_TEs$origin <- "genome"
table_of_cluster_TEs$origin <- "cluster"
clust <- table_of_cluster_TEs[,c(5,1,3)]
genome <- table_of_TEs[,c(5,1,3)]
all <- rbind(clust, genome)
colnames(all)[2] <- "family"

# Stacked barplot
my.stack.out <- paste0("output/18.2-Apul-piRNA-TE-overlap/", my.outprefix,"cluster_stack_plot.pdf")
pdf(my.stack.out)
ggplot(all, aes(fill=family, y=percent, x=origin)) + 
  geom_bar(position="stack", stat="identity") + theme_bw()
dev.off()

```
This figure only shows the proportional enrichment of specific transposable element families respective to other families either in the genome or in the clusters, but does not describe the enrichment of a specific family in piRNA clusters respective to that of the entire genome (% of genome or % of cluster). I need to obtain the fraction of coverage of each family respective to the genome and to the piRNA clusters full lengths


```{r}

# calculate representative fraction of genome (475381253 bases) corresponding to TE families
cov_file.g <- bed_file %>%
  mutate(bases = V7-V6) %>%
  select(V11, bases) %>%
  group_by(V11) %>%
  summarise(total = sum(bases)) %>%
  mutate(perc = total/475381253*100)

# calculate representative fraction of piRNA clusters ( bases) corresponding to TE families
cov_file.c <- cluster_TEs %>%
  select(V8, V9) %>%
  group_by(V8) %>%
  summarise(total = sum(V9)) %>%
  mutate(perc = total/788650*100)  



```


```{r, engine='bash'}
# create a scaffold lengths file "chr" "start(all zeros)" "end(length)"

cat GCF_013753865.1_Amil_v2.1_genomic.fna | awk '$0 ~ ">" {if (NR > 1) {print c;} c=0;printf substr($0,2,100) "\t"; } $0 !~ ">" {c+=length($0);} END { print c; }' | awk 'BEGIN{ FS = OFS = "\t" } { $1 = $1 FS ("0") }1' > Amil.extents.bed


bedmap --echo --bases-uniq --delim '\t' Amil.extents.bed APUL_repeatmasker_bedfile.bed | awk '{ print $0"\t"100*$4/($3 - $2); }' - > Genome_TE_coverage.bed

```

