---
title: "Apul enrichment of miRNA targets"
author: "Jill Ashey"
date: "2024-10-30"
output: html_document
---

## GO enrichment of Apul miRNA targets - using 3'UTRs derived from Amil genome 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(topGO)
```

See [bash code](https://github.com/JillAshey/JillAshey_Putnam_Lab_Notebook/blob/master/_posts/2024-06-15-e5-deepdive-miRNA-TargetPrediction.md) for `miranda` analysis, which was done on Andromeda. 

Read in miranda output file 
```{r}
miranda_1kb <- read.delim("../output/21-Apul-miRNA-target-prediction/miranda_strict_all_1kb_parsed_amil_apul.txt", header = F) 

# renaming columns is being weird, doing it manually
colnames(miranda_1kb) <- c("mirna", "Target", "Score", "Energy_kcal_mol", "Query_start_end", "Subject_start_end", "Length", "Query_Identity", "Subject_Identity")

# Remove > in front of miRNA name 
miranda_1kb$mirna <- gsub(">", "", miranda_1kb$mirna)
length(unique(miranda_1kb$mirna))
length(unique(miranda_1kb$Target))
```

Make gene id column
```{r}
miranda_1kb$gene_id <- sub("^ID=gene-([^;]+);.*", "\\1", miranda_1kb$Target)
```

Create new df with only gene and miRNA info; summarize which miRNAs target which genes 
```{r}
# Create the new data frame with gene_id and concatenated mirnas
gene_mirna_1kb_df <- miranda_1kb %>%
  group_by(gene_id) %>%
  summarize(mirnas = paste(mirna, collapse = ", "))
```

Summarize number of genes targeted by each miRNA
```{r}
mirna_gene_summary <- miranda_1kb %>%
  group_by(mirna) %>%
  summarize(num_genes = n_distinct(gene_id))
```

Read in gene to go information (generated [here](https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/19-Apul-bedtools-closest.Rmd))
```{r}
APUL_gene2go<-read.delim("../../data/Apul_gene2go.tab", sep="\t")
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
apul_clust_genes <- as.character(unique(miranda_1kb$gene_id))

# All genes 
apul_all_genes <- as.character(APUL_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
Apul_GeneList <- factor(as.integer(apul_all_genes %in% apul_clust_genes))
names(Apul_GeneList) <- apul_all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by miRNAs. 

Read in gene-to-go-mappings
```{r}
APUL_gene2go_topgo<-readMappings("../../data/Apul_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_BP <-new("topGOdata", ontology="BP", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_BP_FE <- runTest(GO_APUL_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_BP_En <- GenTable(GO_APUL_BP, Fisher = GO_APUL_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_BP_En$Fisher<-as.numeric(GO_APUL_BP_En$Fisher)
GO_APUL_BP_En_sig<-GO_APUL_BP_En[GO_APUL_BP_En$Fisher<0.05,]
```

Merge `GO_APUL_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_BP_En_sig$GO.ID <- trimws(GO_APUL_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_BP_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_CC <-new("topGOdata", ontology="CC", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_CC_FE <- runTest(GO_APUL_CC, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_CC_En <- GenTable(GO_APUL_CC, Fisher = GO_APUL_CC_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_CC_En$Fisher<-as.numeric(GO_APUL_CC_En$Fisher)
GO_APUL_CC_En_sig<-GO_APUL_CC_En[GO_APUL_CC_En$Fisher<0.05,]
```

Merge `GO_APUL_CC_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_CC_En_sig$GO.ID <- trimws(GO_APUL_CC_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_CC_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_MF <-new("topGOdata", ontology="MF", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_MF_FE <- runTest(GO_APUL_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_MF_En <- GenTable(GO_APUL_MF, Fisher = GO_APUL_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_MF_En$Fisher<-as.numeric(GO_APUL_MF_En$Fisher)
GO_APUL_MF_En_sig<-GO_APUL_MF_En[GO_APUL_MF_En$Fisher<0.05,]
```

Merge `GO_APUL_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_MF_En_sig$GO.ID <- trimws(GO_APUL_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_MF_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_APUL_En_sig_gene <- rbind(GO_APUL_BP_En_sig_gene, GO_APUL_CC_En_sig_gene, GO_APUL_MF_En_sig_gene)
```

Merge `miranda_1kb` with `GO_APUL_En_sig_gene` and calculate proportion of significant genes
```{r}
merged_targets <- GO_APUL_En_sig_gene %>%
  left_join(miranda_1kb, by = c("gene_name" = "gene_id")) %>%
  na.omit() %>%
  mutate(prop.sig.genes = Significant/Annotated)

# Write to csv 
write.csv(merged_targets, "../output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_miRNA_targets.csv")
```

Plot 
```{r}
APUL_GO_plot<-ggplot(merged_targets, aes(x = Term, y = Fisher, size = prop.sig.genes, fill = Fisher)) +
  #expand_limits(y = 1.5) +
  #ylim(1, 7.25) +
  # Add horizontal lines with a single aesthetic value
  #geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  #geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  #labs(caption = 'Cut-off lines at p=0.01 and p=0.001') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); APUL_GO_plot
```

## GO enrichment of peve-mir-100 targets

Interested in looking specifically at miR-100, as this is highly conserved across cnidarians and bilaterians

Before starting this section, run lines 9-57 to get the miranda information. 

Filter by mir-100
```{r}
mir100_miranda_1kb <- miranda_1kb %>%
  filter(mirna == "apul-mir-100")
```

Make list of genes for input to TopGO
```{r}
# Genes of interest - ie those targeted by miRNAs 
apul_clust_genes <- as.character(unique(mir100_miranda_1kb$gene_id))

# All genes 
apul_all_genes <- as.character(APUL_gene2go$gene_name)

# Apply 1 or 0 if gene is gene of interest 
Apul_GeneList <- factor(as.integer(apul_all_genes %in% apul_clust_genes))
names(Apul_GeneList) <- apul_all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test to assess whether specific GO terms are overrepresented in the genes targeted by mir-100. 

Read in gene-to-go-mappings
```{r}
APUL_gene2go_topgo<-readMappings("../../data/Apul_gene2go.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_mir100_BP <-new("topGOdata", ontology="BP", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_mir100_BP_FE <- runTest(GO_APUL_mir100_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_mir100_BP_En <- GenTable(GO_APUL_mir100_BP, Fisher = GO_APUL_mir100_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_mir100_BP_En$Fisher<-as.numeric(GO_APUL_mir100_BP_En$Fisher)
GO_APUL_mir100_BP_En_sig<-GO_APUL_mir100_BP_En[GO_APUL_mir100_BP_En$Fisher<0.05,]
```

Merge `GO_APUL_mir100_BP_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_mir100_BP_En_sig$GO.ID <- trimws(GO_APUL_mir100_BP_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_mir100_BP_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_mir100_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_mir100_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_mir100_CC <-new("topGOdata", ontology="CC", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_mir100_CC_FE <- runTest(GO_APUL_mir100_CC, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_mir100_CC_En <- GenTable(GO_APUL_mir100_CC, Fisher = GO_APUL_mir100_CC_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_mir100_CC_En$Fisher<-as.numeric(GO_APUL_mir100_CC_En$Fisher)
GO_APUL_mir100_CC_En_sig<-GO_APUL_mir100_CC_En[GO_APUL_mir100_CC_En$Fisher<0.05,]
```

Merge `GO_APUL_mir100_CC_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_mir100_CC_En_sig$GO.ID <- trimws(GO_APUL_mir100_CC_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_mir100_CC_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_mir100_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_mir100_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions 

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_APUL_mir100_MF <-new("topGOdata", ontology="MF", gene2GO=APUL_gene2go_topgo, allGenes=Apul_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_APUL_mir100_MF_FE <- runTest(GO_APUL_mir100_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_APUL_mir100_MF_En <- GenTable(GO_APUL_mir100_MF, Fisher = GO_APUL_mir100_MF_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_APUL_mir100_MF_En$Fisher<-as.numeric(GO_APUL_mir100_MF_En$Fisher)
GO_APUL_mir100_MF_En_sig<-GO_APUL_mir100_MF_En[GO_APUL_mir100_MF_En$Fisher<0.05,]
```

Merge `GO_APUL_mir100_MF_En_sig` with GO and gene info. 
```{r}
# Separate GO terms 
APUL_gene2go <- APUL_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
APUL_gene2go$GeneOntologyIDs <- trimws(APUL_gene2go$GeneOntologyIDs)
GO_APUL_mir100_MF_En_sig$GO.ID <- trimws(GO_APUL_mir100_MF_En_sig$GO.ID)

# Join the datasets based on GO term
GO_APUL_mir100_MF_En_sig_gene <- APUL_gene2go %>%
  left_join(GO_APUL_mir100_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_APUL_mir100_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Join ontologies 

Bind so there is a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_APUL_mir100_En_sig_gene <- rbind(GO_APUL_mir100_BP_En_sig_gene, GO_APUL_mir100_CC_En_sig_gene, GO_APUL_mir100_MF_En_sig_gene)
```

Merge `mir100_miranda_1kb` with `GO_APUL_mir100_En_sig_gene` and calculate proportion of significant genes
```{r}
merged_mir100_targets <- GO_APUL_mir100_En_sig_gene %>%
  left_join(mir100_miranda_1kb, by = c("gene_name" = "gene_id")) %>%
  na.omit() %>%
  mutate(prop.sig.genes = Significant/Annotated)

# Write to csv 
write.csv(merged_mir100_targets, "../output/21-Apul-miRNA-target-prediction/Apul_GO_en_sig_mir-100_targets.csv")
```

```{r}
APUL_GO_plot<-ggplot(merged_mir100_targets, aes(x = Term, y = Fisher, size = prop.sig.genes, fill = Fisher)) +
  #expand_limits(y = 1.5) +
  #ylim(1, 7.25) +
  # Add horizontal lines with a single aesthetic value
  #geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  #geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  #labs(caption = 'Cut-off lines at p=0.01 and p=0.001') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); APUL_GO_plot
```