---
title: "miRNA target prediction"
author: "Jill Ashey"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Validating 3'UTR methods 

For Apul, I conducted 2 methods to identify 3'UTRs: 
1) Ran Gene Ext, which extended ~15% of genes. Median extension was 1010bp. For the rest of the genes, I extended them by 1kb on the right flank and subtracted any areas that overlapped with an existing gene. 
2) ONLY extended by 1kb on the right flank and subtracted any areas that overlapped with an existing gene. 

1) miranda results from gene ext + 1kb data 
```{r}
# Read in data 
miranda_all <- read.delim("../output/21-Apul-miRNA-target-prediction/miranda_strict_all_3UTRs_parsed_Apul.txt", header = F) %>% 
  rename(mirna = V1,
        Target = V2, 
        Score = V3, 
        Energy_kcal_mol = V4, 
        Query_start_end = V5, 
        Subject_start_end = V6,
        Length = V7, 
        Subject_Identity = V8, 
        Query_Identity = V9)

miranda_all$mirna <- gsub(">", "", miranda_all$mirna)
length(unique(miranda_all$mirna))
length(unique(miranda_all$Target))

# Select energy <-20, make new gene id column and remove all coord and length info 
miranda_filt <- miranda_all %>%
  filter(Energy_kcal_mol < -20) %>%
  separate(Target, into = c("gene_id", "Target_coords"), sep = "::") %>%
  select(-c(Target_coords, Query_start_end, Subject_start_end, Length)) %>%
  unique()
length(unique(miranda_filt$mirna))
length(unique(miranda_filt$gene_id))

# Create the new data frame with gene_id and concatenated mirnas
gene_mirna_df <- miranda_filt %>%
  group_by(gene_id) %>%
  summarize(mirnas = paste(mirna, collapse = ", "))

# Create the new data frame with the highest score mirna for each gene_id
gene_top_mirna_df <- miranda_filt %>%
  group_by(gene_id) %>%
  slice_max(Score, n = 1, with_ties = FALSE) %>%
  select(gene_id, mirna, Score)
```

2) miranda results from 1kb data only 
```{r}
# Read in data 
miranda_1kb <- read.delim("../output/21-Apul-miRNA-target-prediction/miranda_strict_all_1kb_parsed_apul.txt", header = F) %>% 
  rename(mirna = V1,
        Target = V2, 
        Score = V3, 
        Energy_kcal_mol = V4, 
        Query_start_end = V5, 
        Subject_start_end = V6,
        Length = V7, 
        Subject_Identity = V8, 
        Query_Identity = V9)

miranda_1kb$mirna <- gsub(">", "", miranda_1kb$mirna)
length(unique(miranda_1kb$mirna))
length(unique(miranda_1kb$Target))

# Select energy <-20, make new gene id column and remove all coord and length info 
miranda_1kb_filt <- miranda_1kb %>%
  filter(Energy_kcal_mol < -20) %>%
  separate(Target, into = c("gene_id", "Target_coords"), sep = "::") %>%
  select(-c(Target_coords, Query_start_end, Subject_start_end, Length)) %>%
  unique()
length(unique(miranda_1kb_filt$mirna))
length(unique(miranda_1kb_filt$gene_id))

# Create the new data frame with gene_id and concatenated mirnas
gene_mirna_1kb_df <- miranda_1kb_filt %>%
  group_by(gene_id) %>%
  summarize(mirnas = paste(mirna, collapse = ", "))

# Create the new data frame with the highest score mirna for each gene_id
gene_top_mirna_1kb_df <- miranda_1kb_filt %>%
  group_by(gene_id) %>%
  slice_max(Score, n = 1, with_ties = FALSE) %>%
  select(gene_id, mirna, Score)
```

Summarize number of genes targeted by miRNAs via each method 
```{r}
# Gene ext + 1kb
mirna_gene_summary <- gene_top_mirna_df %>%
  group_by(mirna) %>%
  summarize(num_genes = n_distinct(gene_id))

# 1kb 
mirna_gene_1kb_summary <- gene_top_mirna_1kb_df %>%
  group_by(mirna) %>%
  summarize(num_genes = n_distinct(gene_id))

# join dfs to compare numbers 
both_methods <- full_join(mirna_gene_summary, mirna_gene_1kb_summary, by = "mirna")
```

Both methods identify similar numbers of target genes, which is a good check. Are they targeting the same genes? 
```{r}
merged_df <- full_join(miranda_filt, miranda_1kb_filt, by = c("mirna", "gene_id"))

# Count NAs in `Score.x` and `Score.y`
na_count_x <- sum(is.na(merged_df$Score.x))  # Rows present only in `miranda_filt` - gene ext + 1kb
na_count_y <- sum(is.na(merged_df$Score.y))  # Rows present only in `miranda_1kb_filt` - 1kb only 

# Count rows that matched in both data frames (non-NA in both `Score.x` and `Score.y`)
match_count <- sum(!is.na(merged_df$Score.x) & !is.na(merged_df$Score.y))

# Display results
na_count_x
na_count_y
match_count
```

There are 5820 interactions that are identified in both! <300 interactions are only in one method or the other. The methods are yielding very similar results, which is a good gut check. I think I will go with the 1kb method for now, as it does not require bam files, whereas gene ext needs bam files to perform extensions. 

Also note that the interactions described here are not necessarily the ones that are active in the coral at this particular time. This analysis only tells us the potential mRNA targets of miRNAs. Expression data is needed to understand regulation dynamics. 


